---
title: "ADOPT: CODING STYLE GUIDE"
author: "Strategic Data Project"
date: "Center for Education Policy Research at Harvard University"
output:
  pdf_document:
    latex_engine: xelatex
    toc: no
    toc_depth: 4
    includes:
      in_header: harvardheader.tex
      before_body: harvard_prefix.tex
  html_document: default
---


# ADOPT: CODING STYLE GUIDE

## Introduction

### Overview

In our context, ‘Programs' are coded instructions to conduct analyses using
statistical software. These programs are often co-written by several analysts
working collaboratively. Since most analysts work in teams, programs written to
conduct statistical analyses serve as building blocks for increased knowledge
sharing across a department. These programs, when written clearly, can be
sampled or entirely reused by other analysts to avoid “reinventing the wheel.” 

The Center for Education Policy Research (CEPR) and the Strategic Data Project
(SDP) recommend analysts follow published standards when writing programs.
Though every programmer has her own idiosyncratic style, CEPR encourages using
standards that facilitate a common methodology for creating programs and sharing
code. 

The CEPR Coding Style Guide is intended for analysts familiar with
statistical programs that require some level of programming. Ideally, analysts
who use this guide have manipulated large datasets and conducted statistical
analyses with a programming language such as Stata or SQL. If you have not used
these software programs, but are comfortable writing code, this Coding Style
Guide will be accessible to you.

### Scope

The CEPR Coding Style Guide establishes standards for writing programs and codes
to conduct statistical analyses in Stata – the software CEPR's research team
uses. However, many conventions, are applicable to any programming language.
These conventions are meant to facilitate standardization, not provide insight
into language functionality or syntax.

This guide outlines best practices to:

- name data files and folders
- name variables and macros within a program
- establish code structure
- document and comment on code
- write programs to ensure understanding across a team of analysts

The variables and file names used serve only as examples for syntax.

### Document Structure

This document is broken into three sections:

- Naming Conventions
- Commenting and Readability
- Coding Guidelines

The latter two sections contain examples of code in either green or red blocks.
Green blocks are exemplar code that should be mimicked, red blocks are coding
style that should be avoided.

### Terminology

As mentioned before, nomenclature used in this document is based on Stata
terminology. Please use the table below to guide your understing of key terms in
this document:

R Term        | Stata Term    | SQL Term      | C# Term
------------- | ------------ | -------------  | ------------
Row           | Observation  | Row or record  | Object
Column        | Variable     | Column         | Field or attribute
Variable      | Macro        | Variable       | Variable


## Naming Conventions

### General Naming Conventions

When you name folders, files, variables, or macros, keep in mind that the name
should be intuitive enough for others to interpret the meaning and content at
first glance. In general, do not shy away from longer descriptive names with
multiple words as opposed to shorter abbreviated names that may not convey the
proper meaning. Other general guidelines for naming:

- Separate words in any file, folder, or variable name with underscores (_) to 
ensure readability. Avoid using spaces in names.
- Avoid using other punctuation such as periods, hyphens, exclamation points, etc.
- Another alternative is camelCase, which can be preferable in some environments

### Abbreviations and Acronyms

To ensure that names are not overly lengthy, the following abbreviations are 
accepted for commonly used terms:

Term          | Abbreviation
------------- | -------------
School        | sch
Principal     | prn or p
Student       | stu or s
Teacher       | tch or t
Class         | cls or c
Grade Level   | gr
Subject       | subj
Year          | yr
Primary       | pri
Secondary     | sec
College       | clg
Graduate/Graduation | grad
Elementary School | es
Middle school | ms
Junior High   | jhs
High School   | hs
Charter School| chs
Maximum       | max
Minimum       | min
Mean or Average | avg
Count         | cnt
Date          | dt
Number        | num or n
Standard Deviation | sd
English Language Arts | ela
Verbal        | verb
Performance   | perf

Additional terms can be abbreviated only if the abbreviation is intuitive. When 
in doubt, spell it out! 

Also, it is acceptable to use acronyms for partner agencies, projects, and analyses. 

For example:

Term              | Abbreviation
----------------- | -------------
Strategic Data Project | SDP
National Center for Teacher Effectiveness | NCTE
The New Teacher Project | TNTP
Fort Worth Independent School District | FWISD
Delaware Department of Education | DEDOE
Human Capital Diagnostic | HK
College Going Diagnostic | CG
Strategic Performance Indicator | SPI

### Folder Naming and Structure

To organize files properly on any server, we recommend you adhere to the
following guidelines:

- The root folder should be named according to the project that data and
analyses contained within belong to (e.g. School_Reports). - Sub-folders within
a project should be named according to their contents, with the most common
being "data" (with sub-folders for "raw" and "clean" files), "programs" or
"do_files", "logs" or "log_files", "notes", and "tables_figures".

For those programming in a Windows environment, pin commonly accessed folders to
the Favorites bar in Windows Explorer for easier access.

### File Naming

Files should be named using "compositional identifiers" that allow an individual
to understand the contents of a file at a high level without having to open the
file. This is especially important for program files (e.g. R .R files or
SQL scripts) and graphs. The Compositional identifiers file name should descend
in order of importance so the files group together in an intuitive order when
sorted by name (default in Windows Explorer). For example, files should be named
with the following compositional identifiers in the following order of
importance:

- Project Name,
- Component of process - for instance student demographic data or survey data,
- Date (in YYYYMMDD format), 
- version number, or 
- state of file (i.e. temp, test, review, final, etc)

So, for example:

- School_Report_Student_Attributes_20110601_DRAFT.do
- School_Report_Student_Attributes_20110601_REVIEW.do
- School_Report_Student_Attributes_20110601_FINAL.do

Even though folder structure may imply the contents of a file and the above
guidelines may seem redundant, files can be shared across departments in your
organization or with other external entities and therefore names should convey
the same meaning outside of folder structure.

Additionally, file names should be as consistent as possible, especially output
files (graphs, logs) related to a program file. For example, a graph output of
the above R .R file may be named
`School_Report_Student_Attributes_20110601_FINAL_ethnicity.png`.


### Variable Naming

The number of characters used to name variables is limited. Database columns may
be limited to 30 characters depending on platform. Additionally, many R commands
only print a few characters by default. Keep this in mind when you name
variables. Try to be both specific and concise in your variable names.

All variables in an R dataframe should be named (as should database columns). 
R does not have functionality to add additional labels to describe the data. 
These additional variable descriptions should be listed in a separate .R file 
that you can call in your main code, in a list object that stores information 
about key variables, or externally in a common external file. 

```{r variableNaming, eval=FALSE}
# label variables in standard student file
source("programs/dcps_student_labeling.R")

# label variables not in a standard labeling file
varList <- list(stu_struc_move = "student had a structural move, moving schools")

```


## Commenting and Readability

### Comments

Comments are important in any program (R, SQL, or otherwise) and should be used
generously but also deliberately. Comments should be used as often as necessary
to explain the logic and use of each portion of code without being burdensome to
write or cluttering the script file. The goal of commenting is to give insight
into your program – not only for others (e.g. your peers) but also for you! When
you've spent hours, days, weeks, months, or years between writing and reading
your own code, you'll thank yourself for the comments!

Generally speaking, a comment every few lines of code is good practice. Try to
write comments that address one or both of these questions:

- What does this block do?
- Why did I implement this block this particular way?

### General Commenting Guidelines

Comments can be entered in R, SQL, and other programming languages using a
number of different syntaxes. Some of the syntaxes are common to most (or all)
languages, some not. To ensure consistency between types of programs (R,
SQL, etc), please use the following guidelines. These guidelines ensure that
your programs are readable when printed or viewed in a normal text editor that
does not do syntax highlighting.

- Use the # for commenting in R either at the beginning of a line or as an 
end-of-line comment (a comment at the end of a line of code)

```{r commentExample2, eval=FALSE}
# Comment explaining something
table(myvariable) # end-of-line comment

```

- Use commented lines of -- to set off blocks of comments and to break up 
script files
- Precede each line of the comment block with a single comment character.

```{r commentExample}
# Block one ---------
# This is some
# Additional things to explain
# ---------------
```

- In R you cannot comment within a line of code

```{r inlineBlocks, eval=FALSE}

data$variable <- # comment # value

```

- Keep comments as succinct as possible (e.g., one line) while not losing meaning.
- Leave one space between the # and your comment's first character.

```{r goodComment, eval=FALSE}
# compute average test score
test_avg = mean(score)
```

Not:

```{r badComment, eval=FALSE}
# The following line is meant to compute the average of a student's test score.
test_avg = mean(score)
```

- If a single-line comment needs to be long enough to extend beyond the
screen/page width, turn it into a block comment.

- Similarly, if a single line of code (command) is long enough to extend beyond
the screen/page, break the code into multiple lines. Always indent the 
continuing lines of code. Many code editors will do this for you.

```{r groupedValues, eval=FALSE}
library(dplyr) # load needed package

grouped_data <- data %>% group_by(s_year, schoolname) %>% 
  summarize_all(.funs = "mean")

```

- Use a string of dashes in a comment to distinguish between a high-level
description of a block of code from more granular comments. End the block with a
similar string.

```{r globals, eval=FALSE}
# ------------ Data Set Globals -----------------
location = "C:\test"
file_name = "file1"
source_id = "studentid"
xwalk_file = "id_xwalk"
masked_id = "sdpsid" # Variable containing masked id in crosswalk
# ------------ Data Set Globals -----------------
```

- Mark the end of a large block of nested conditional logic – such as a loop or
if statement.

```{r loopFormat, eval=FALSE}
for(subj in c("math", "read")){
  for(yr in c(2001:2005)){
    tmp_yr <- yr + 4
#   / LOTS OF STUFF
#    ...
#    ...
#    ...
    } # End of loop over years
 #// MORE STUFF
# ...
# ...
  } # End of loop over subjects

```

- End-of-line comments are discouraged (except for annotating a group of aligned
variables or marking the end of a block of nested code). Instead, comments
should precede a line of code.

```{r mergeComment, eval=FALSE}
# Merge teacher VA math
full_data <- inner_join(student_data, school_data, 
                        by = c("studentid", "year"))
```

Not:

```{r mergeCommentBad, eval=FALSE}
full_data <- inner_join(student_data, school_data, 
                        by = c("studentid", "year")) # Merge teacher VA math
```

- Commenting can help you keep track of future analyses you'll want to come back
to. For example, make explicit call-outs to pieces of code that need to be
implemented, reviewed, updated, or configured with "TO DO".

```{r}
# TODO: generate new hire and late hire variable using hire date
# TODO: update variables based on school year
# TODO: test the following block for correctness on a larger data set
```

### File Headers

At the top of program files should be a block comment that summarizes your
program, names the file, names the author (first letter of first name and last
name), lists the date created, provides a description, and lists inputs,outputs,
and updates. The description should walk through high level logical steps. These
steps should be identified in the actual code. Consider what is done here:

```{r fileBlock, eval=FALSE}
###################################################################
# File name: crosswalk_masked_ids.do
# Author(s): JSilver
# Date: 5/27/11
# Description: This program creates the crosswalk of student ids to random
#               research ids by:
#  1. Inputting the universe of student ids
#  2. Filtering the distinct set of student ids
#  3. Generating random ids and associating to student ids
# 
# Inputs: ../raw/students/studentyearsch.dta
#          ../raw/students/englang.dta
# 
# Outputs: ../data/bps_student_school_year.dta
# 
# Update 1: TKawakita, 6/1/11 - Added check to ensure random ids are unique
#############################################################################
options(set.width = 80) # set console output width
# Declare any global varaibles
RAWDIR <- "cepr-files/projects/DCPS/Raw"
DATADIR <- "cepr-files/projects/DCPS/Data"
LOGIDR <- "cepr-files/projects/DCPS/Log Files"

#- Step 1: Input universe of student ids ##
...
#- Step 2: Filter distinct set of student ids ##
...
#- Step 3: Generate random ids and associate to student ids ##
...
#- Update 1: Add check to ensure ids unique ##
...

```

### White Space and Readability

White space refers to indentation and extra lines that make code readable. Lack
of white space is referred to as "spaghetti code" since it is difficult to
ascertain where one command ends and the next begins.

Code should be indented to make clear which blocks of code are nested inside of
others (especially when working with loops or conditional statements). An indent
should result in four spaces so that code prints and displays identically within
other applications or computers.

The following is an example of well indented code:

```{r indentExample, eval=FALSE}
for(g in c(10, 11, 12)){
  stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == g] <-
  stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == g] - (g - 9)
}
```

The following is an example of the same code poorly indented:

```{r badIndent, eval=FALSE}
for(g in c(10, 11, 12)){
stusy$temp4_first9year[stusy$grade_flag_max == 1 & stusy$first_9th_flag == 1 & 
stusy$grade_level == g] <- stusy$school_year[stusy$grade_flag_max == 1 & 
stusy$first_9th_flag == 1 & stusy$grade_level == g] - (g - 9)
}
```

An opening curly brace should never go on its own line and should always be
followed by a new line. A closing curly brace should always go on its own line,
unless it’s followed by else.

Always indent the code inside curly braces.

```{r goodBrace, eval=FALSE}
if (y == 0) {
  log(x)
} else {
  y ^ x
}

```

Not:

```{r badBrace, eval=FALSE}
if (y == 0) 
{
  log(x)
} 
else 
{
  y ^ x
}
```

As mentioned earlier, if a single command or line of code extends beyond the
80 characters, break the command into several lines, indent the lines following
the start of the command.

```{r breakGroupBy, eval=FALSE}

progressBars <- bind_rows(
  # Data set one
  plotdf %>% group_by(ot) %>% tally() %>% ungroup %>% 
    mutate(count = sum(n), first_hs_name = "Agency Average"),
  # Data set two
  plotdf %>% group_by(first_hs_name, ot) %>% tally() %>% ungroup %>% 
    group_by(first_hs_name) %>%
    mutate(count = sum(n))
)

```


Not:

```{r breakBadGroup, eval=FALSE}
progressBars <- bind_rows(plotdf %>% group_by(ot) %>% tally() %>% ungroup %>%     mutate(count = sum(n), first_hs_name = "Agency Average"), plotdf %>% group_by(first_hs_name, ot) %>% tally() %>% ungroup %>% 
group_by(first_hs_name) %>% mutate(count = sum(n)))

```

Follow the same guidelines for R graphing code:

```{r graphCode, eval=FALSE}
ggplot(progressTrack, aes(x = outcome, y = measure, group = ot,
           color = ot, linetype = ot)) + 
  geom_line(size = 1.1) + 
  geom_point(aes(group = 1), color = I("black")) + 
  geom_text(aes(label = round(measure * 100, 1)), vjust = -0.8, 
            hjust = -0.25, color = I("black")) +
  geom_text(data = ann_txt, aes(label = textlabel)) + 
  scale_y_continuous(limits = c(0, 1), label = percent) + 
  theme_bw() + theme(legend.position = c(0.825, 0.825)) + 
  scale_color_brewer(type = "qual", palette = 2) +
  guides(color = "none", linetype = "none")

```

When multiple commands with the same function are grouped together, they can
be properly indented and the components of the command should be aligned.

```{r cleanBlock, eval=FALSE}
loc             =  "C:\test"  # Location of files
file_name       = "filename1" # Name of source data file
source_id       = "studentid" # Variable containing the unique id in data file
xwalk_file_name = "id_xwalk"  # Name of crosswalk file
masked_id       = "sdpsid"    # Variable containing masked id in crosswalk
```

Not:

```{r badBlock, eval=FALSE}
loc = "C:\test"  # Location of files
file_name = "filename1" # Name of source data file
source_id = "studentid" # Variable containing the unique id in data file
xwalk_file_name = "id_xwalk"  # Name of crosswalk file
masked_id = "sdpsid"  # Variable containing masked id in crosswalk
```

Finally, more white space is better than less. Make it easy for peers to read
your code!

## CODING GUIDELINES

### Initializing Your Environment (R)

The first guideline when writing an R program, either within a .R script or
interactively, is to initialize your environment. This is done by: 

1. Loading any external packages you will need
2. Setting any environment options
3. Declaring global variables

These should be the first commands following your file header.

```{r envExample, eval=FALSE}
# Initialize packages
library(dplyr)
library(ggplot2)

# Set options
set.options(width = 80)

# Delcare global variables

MODE <- "PRODUCTION"
```

### Logging Output (R)

It is important to log the output of R programs and ad-hoc commands so the
results can be reviewed later. R offers two options for logging your output. 
The first is a traditional log of input and output familiar in other programming 
languages, the second is the concept of an interactive notebook which combines 
R code, markup, and output in one interactive document. 

#### Logs

R does not automatically capture output displayed to a log file. If a log file
is not explicitly opened before commands are made, the results can only be saved
if they are copied out of R's output window. Log your work as a text file so it
can be viewed outside of R:

```{r sinkExample2, eval=FALSE}
sink(file = "log1.txt")
```

The log must be closed at the end of your program/work. Otherwise it will not be saved!

```{r sinkClose, eval=FALSE}
# Close sink
sink()
```

Capture segments of your .R file to separate logs rather than having one log
for the entire program. For instance, rather than having one log file for the
entire School Report analysis, you may want to log each component of the
analysis. Survey analyses would be logged separately from student demographics
analyses.

```{r sinkExample, eval=FALSE}
sink()
rm(matrix); gc()

if (new_hires_by_pov == 1) {
  sink(file = "log/dcps_recruitment_new_hires_by_pov.txt")
  # OTHER STUFF
  sink()
}
rm(matrix); gc()
if (late_hire_over_time == 1) {
  sink(file = "log/dcps_recruitment_late_hire_over_time.txt")
  # OTHER STUFF
  sink()
}

```

#### Notebooks

Using RStudio or other R IDEs it is possible to create R Markdown Notebooks. 
To learn more about these, [read about them from RStudio.](http://rmarkdown.rstudio.com/r_notebooks.html)

### Global Variables as Switches

You may want to run only parts of the program at a time. To do this, use global 
variables that act as "switches" to section off distinct and independent parts 
of the program. A switch is essentially a variable turns parts of your program
on and off. In R it is traditional to code these as logical.

```{r globalVarExample, eval=FALSE}
# SWITCHES
teacher <- FALSE
student <- FALSE
test <- FALSE
. . .
if(teacher){
  # Some code run if teacher = TRUE
} 
if(student) {
  # Some code run if student = TRUE
}
if(test) {
  # Some code run if test = TRUE
}
```

You may also put brackets around code without a condition or global. In most R 
text editors, this will allow you to collapse any code in the brackets and 
minimize the amount of scrolling up and down you will need to do. 

```{r bracesCollapse, eval=FALSE}
{
# Code to collapse
}
```

### Conditions

Conditions, or conditional code, refer to pieces of code that execute if a
certain Boolean logic statement is "true". Conditions take the form of if/else
statements or loops. Conditional logic was used in the previous example to
demonstrate switches.

A code block executed as a result of a condition should be encapsulated in
braces ({ }). The open-brace or left-brace ( { ) should always be on the same
line as the condition and the close-brace or right-brace ( } ) should always be
on its own line:

```{r eval=FALSE}
if(x > 0) {
  print("x is positive")
}
else if (x < 0) {
  print("x is negative")
}
```

Not:

```{r eval=FALSE}
# First bad example
if x > 0
{
print("x is positive")
}
# Second bad example
else if x < 0
{print("x is positive")}
```

If a branch involves a single statement, then braces are not necessary. However,
it is still advised to use them in case more statements are added within the
condition later. Regardless, always put the conditional code on the line
following the condition and never on the same line as the condition:

```{r eval=FALSE}
if x > 0
  print("x is positive")
else if x < 0
  print("x is negative")
```

Not:

```{r badConditional, eval=FALSE}
if x > 0 print("x is positive")
else if x < 0 print("x is negative")
```

### Hard Coding vs Macros

Hard coding is the practice of using literal values in code instead of
variables. Hard coding should be avoided whenever possible. Take the following
code (which generates an expected school year for a student in grade 9 as 
an example:

```{r eval=FALSE}
# Grade 10
stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 10] <-
  stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 10] - 1
# Grade 11
stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 11] <-
  stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 11] - 2
# Grade 12
stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 12] <-
  stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 12] - 3
```

Instead of hard-coding grades and numbers of years, variables and looping can 
be used:

```{r eval=FALSE}
for(g in c(10, 11, 12)){
  stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == g] <-
  stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == g] - (g - 9)
}
```

Initially, it may seem that hard coding is more intuitive and easier to read.
However, hard-coding is much more difficult to maintain. Take the scenario where
the two sections of code above were to be changed to function on different grade 
levels or with different filters. For the hard-coded program,
many literal values would need to be changed. In the more elegant version that
uses macros defined at the beginning of the program, only a few changes need 
to be made.

If hard coding appears necessary (though with some more thought it likely
is not), make a large and distinct call out to this with a comment.

### Macros as File Paths

When defining file paths using macros to define "input" information is
preferable to hard-coding. Often, a program references input data or output
locations and files. By using a global macro at the beginning of the program,
you can easily change the location of input and output files. This is especially
important when transporting your program outside of your work environment to
work on-site.

```{r eval=FALSE}
# Location of input data files
location <- "C:\test\data\input"
# List of data file names
filenames <- c("filename1", "filename2", "filename3")
# ----------------------------------
# Change directories to location of files
setwd(location)
#- Merge masked ids onto files
for(filename in filenames){
  load(filename)
}
```


### Closing

This guide is not an exhaustive list of best practices for programming. However,
these guidelines will equip you and your team to conduct analyses effectively,
efficiently, and reliably. These coding practices, will improve your ability to
share knowledge across your organization and build capacity to construct
sophisticated analyses with statistical software.


\vfill
Last updated January 2017.

© 2017 President and Fellows of Harvard College. 