---
title: "SDP Data Building Tasks"
author: "Strategic Data Project"
date: "October 31, 2016"
output: 
  html_document: 
    toc: yes
---


# SDP Data Building Tasks


```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, comment=NA}
# Set options for knitr
library(knitr)
knitr::opts_chunk$set(comment=NA, warning=FALSE, 
                      error=FALSE, message=FALSE, fig.align='center')
options(width=120)
```

## Overview

Housed at the Center for Education Policy Research at Harvard University, the 
Strategic Data Project (SDP) partners with school districts, school networks, 
and state agencies across the US. **Our mission is to transform the use of 
data in education to improve student achievement.** We believe that with the 
right people, the right data, and the right analyses, we can significantly 
improve the quality of strategic policy and management decisions. 

### Core Strategies
To achieve our mission, SDP pursues three core strategies: 

1. Placing top-notch analytic leaders as ``Fellows'' for two years with our 
partner agencies; 
\noindent{\small SDP supports more than 40 Data and Agency Fellows serving 
partner educational agencies -- districts, states, and charter management 
organizations--across the nation.  This number will grow to nearly 70 in 2012.} 

2. Conducting rigorous diagnostic analyses of teacher effectiveness and 
college-going success using existing agency data; and 

\noindent{\small We have completed diagnostics in teacher effectiveness 
and / or college-going success in seven districts, with more diagnostics 
currently underway or planned in additional district 
and state partner agencies.} 

3. Disseminating our tools, methods, and lessons learned to many more education 
agencies. 

\noindent{\small Through the diagnostic analyses, we have developed a body of 
knowledge around effective data use.  The release of this toolkit reflects SDP's 
third core strategy to spread knowledge and build capacity within educational 
agencies for effective data use.} 


### SDP DIAGNOSTICS

Our second core strategy, conducting rigorous diagnostic analyses using existing 
agency data, focuses on two core areas: **(1) college-going success and 
attainment for students and (2) human capital** (primarily examining teacher 
effectiveness).

The diagnostics are a set of analyses that frame actionable questions for 
education leaders.  By asking questions such as, 
*How well do students transition to postsecondary education?* or 
*How successfully is an agency recruiting effective teachers?* we support 
education leaders to develop a deep understanding of student achievement in 
their agency. In an effort to make these analyses accessible and more widely 
used, this toolkit helps analysts collect data and produce analyses associated 
with the SDP College-Going and Human Capital diagnostics. 

Notably, the diagnostic analyses in this release of our toolkit are specific to 
the College-Going diagnostic.  The data collection (Identify), data cleaning 
(Clean), and best practices (Adopt) stages of the toolkit, however, are 
applicable to either diagnostic and convey general data use guidelines valuable 
to any analysts interested in increasing the quality and rigor of their analyses. 
Later releases will address the analyses in our Human Capital diagnostic.

## Introduction
### SDP Data Building Tasks

\normalsize
Congratulations on identifying the data elements that are essential for 
conducting rigorous analyses in your organization. **Clean** is the next stage 
in the SDP Toolkit for Effective Data Use.  To successfully move through the 
**Clean** stage, you should review the  **Identify** component of this toolkit.  
Upon completing this stage, you will have produced clean research files that 
will allow you to **Connect** and **Analyze** data related to college-going 
success in your agency.

### THE TASKS

**Clean** consist of five tasks that share a similar structure.  The tasks are 
geared toward analysts with at least moderately strong data background and 
comfort with statistics.  Each task provides hands-on experience building 
specific components of the research file used for the SDP CollegeGoing 
Diagnostic Analyses.

The tasks are listed as follows:
- Task 1 Student Attributes
- Task 2 Student School Year
- Task 3 Identifying the Ninth Grade Cohort
- Task 4 Student School Enrollment
- Task 5 Prior Achievement

Each task is accompanied by a practice file dataset upon which all data 
snapshots and output are based. These datasets consist of simulated data that 
have been fully de-identified.We strongly recommend that you use these datasets 
to work through the tasks and check your answers.  The datasets are available 
for download at (www.gse.harvard.edu/sdp/tools)[www.gse.harvard.edu/sdp/tools]. 
Note that the tasks follow a logical sequence from Task 1 to Task 5, and some 
tasks require the output of previous tasks.  However, because we provide all 
necessary practice files for each task, you may also choose to work on the tasks 
out of order.  For instance, you may be first interested in identifying the 
ninth-grade cohort for students in your agency with Task 3. 

To successfully complete all parts of this toolkit, however, you should work 
your way through all five tasks.  The output of each task will be needed to 
successfully complete the Connect and Analyze stages of the toolkit. Lastly, it 
is important to note that the tasks do not show you how to develop every single 
component and detail of the files to be used in Connect and Analyze.  Our goal 
is to equip you with an understanding for the core process of constructing 
robust, clean research files.  We do, however, aim to explicitly indicate what 
additional elements are needed in the `DATA DESCRIPTION` section of each task to 
deliver a fully realized research file.  Furthermore, we also provide a 
`DECISION RULES GLOSSARY` in the Appendix at the end of this document to provide 
guidance on how to approach the cleaning process for these additional elements.

For those who are less familiar with or who need to brush up on Stata use, we 
also include a `R GLOSSARY` of commonly used commands in the Appendix at the end 
of this document.  Through this set of tasks, you will learn effective practices
for: data transformations, new variable construction, and the implementation of 
key decision rules.

### TASK STRUCTURE

The core of each task is a set of step-by-step instructions that guide you 
through the work. For each task you will find:

- Purpose --- Clarifies the importance of the task.  
- How to Start --- Identifies the input file(s) you will need to complete the 
task and guidelines for apply the task to your  own agency's data.
- Data Description --- Lists the data elements you will need to complete the 
task and describes the uniqueness of key data elements.
- Instructions --- Provides logical instructions on transforming the data with 
R code and fill-in-the-blank snapshots that help you visualize changes to 
your data.
- Solutions --- Provides answers for the data snapshot exercises.

After completing these tasks, you will be well-positioned to use your own 
agency’s data to construct similar clean research files needed in the Connect 
and Analyze stages.

Finally, if you find yourself in need of additional guidance, the friendly 
research team at SDP is available to help: `sdp@gse.harvard.edu`

## Task 1: STUDENT ATTRIBUTES
### PURPOSE

Through `Task 1: Student Attributes`, you will take the raw Student Attributes 
file and generate a cleaned Student Attributes output file that has only one 
observation per student. These data will allow you to examine college-going 
outcomes by race/ethnicity.

The core assignments of this task are to:

1. Resolve instances in which the same student appears with different values for 
race/ethnicity in different years. Our goal is to have only one race/ethnicity 
associated with each student.

2. Drop duplicate observations so the file is unique by student--that is, it 
contains only one observation per student. Upon completing this task, you will 
be have a clean Student_Attributes file that can then be used as to create the 
analysis file in Connect.  From Task 1, Task 2 is a natural next step, in which 
you will clean the Student_School_Year file in preparation for Task 3 and 
Task 4.

### HOW TO START
To begin, open the provided Student_Attributes practice file.

```{r echo=TRUE}
# Read in Stata
library(haven) # required for .dta files

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_Demographics_Raw.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stuatt <- read_stata(con) # read data in the data subdirectory
```

The input file contains data for school years 2000-01 through 2006-07.  
Normally race is considered a time-invariant variable that is unique by student.  
In this instance, we deal with a case in which race is stored in a file unique 
by student and school year, which is instead time-variant.  This task aims to 
take convert the dataset from being time-variant to being time-invariant. 

If this is your first time going through the task, we recommend starting with 
the practice file, rather than your agency's own data file.  Doing so will help 
you learn SDP's cleaning methodology and allow you to easily check your answers 
from a common dataset.  You may then apply these methods to your agency's own 
`Student_Attributes` data with confidence.  To learn more about the data you 
will need to collect in your agency, refer to `Identify: Data Specification Guide` 
and the `DATA DESCRIPTION` section of this document.

In addition to the practice file, you may also find it useful to complete the 
data snapshot exercises provided in the task.  These exercises will allow you to 
visualize changes to the data occurring in each step of the task.  Solutions for 
the exercises are provided at the end of the task. 

### DATA DESCRIPTION

In Identify: Data Specification Guide, we specify the data elements included in 
the `Student_Attributes` research file.[^You may be wondering how  this specification 
compares to the  version in Identify: Data Specification Guide. Here are the 
primary changes: First, the `race_ethnicity` variable is coded as a string rather 
than being numeric, as specified in the Data Specification Guide.  You will 
correct this in the task as it will facilitate the process of making the file 
unique by sid.Second, we are examining a time-variant data set. In the Data 
Specification Guide, the `Student_Attributes` file is specified as being unique 
by sid.  In this case, the data are time-variant and unique by sid and 
`school_year`. Note that some districts may actually store `race_ethnicity` in a 
time-variant form such as this, and it is our job through this task to make the 
data time-invariant, i.e. each student only has a single value for `race_ethnicity` 
across time. Third, we are examining a partial data set including only `sid`, 
`school_year`, and `race_ethnicity`.  We do not include variables such as `male`, 
`hs_diploma`, or `hs_diploma_type`, or `hs_diploma_date` to simplify the task.  
These variables are essential for later analyses but are left for you to 
complete as a further exercise.For guidance on cleaning these additional 
variables, refer to the DECISION RULES GLOSSARY at the end of this document and 
use this task as a reference.]

In this task, we examine a partial version of the `Student_Attributes` file that 
includes only `sid`, `school_year`, and `race_ethnicity`. This partial version is 
presented to help you learn the `Student_Attributes` cleaning process to make a 
file unique by sid without having to worry about additional `Student_Attributes` 
variables such as `male`, `hs_diploma`, `hs_diploma_type`, or `hs_diploma_date`.  
The relevant variables and definitions you will need to complete the task are 
illustrated below:

```{r echo=TRUE}
dplyr::glimpse(stuatt)
```

**Uniqueness:** ideally, the data in its raw form would be unique by `sid`.  
However, this may not be the case as some agencies might record `race_ethnicity` 
in a time-variant manner, such as by school year. To address this, we explain 
how to take the raw research file from being unique by `sid` and `school_year` 
to being unique by `sid` alone.  Once the file is unique by `sid` alone, it is 
ready to be incorporated into the analysis file in the Connect stage.

Examine  your `Student_Attributes` raw research file input dataset.  According to 
the data specification, the file should be unique by `sid`.  Examine the snapshot 
below to determine if it is unique as described.  

```{r echo=TRUE}
head(stuatt)
# A quick way to test this in R
length(unique(stuatt$sid)) == nrow(stuatt)
```

```{r note1, include=FALSE}
# Optional--consider removing since R does not need factors to be recoded as 
# numeric. Recode the raw race\_ethnicity variable as numeric. Race\_ethnicity is 
# currently coded as a string variable, which is how some  agencies may store 
# this data . Replace the string values with numeric values as shown below. 
# This numeric race variable will be easier to use in later stages of the task. 
```

### Step 0: Read in the Data

```{r echo=TRUE}
# Read in Stata
library(haven) # required for .dta files

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_Demographics_Raw.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stuatt <- read_stata(con) # read data in the data subdirectory
stuatt <- as.data.frame(stuatt)
```


Now drop the `first_9th_school_year_reported` variable. You will create a 
`first_9th_school_year_reported` variable in Task 3 that also imputes this 
variable for transfer-ins. 

```{r echo=TRUE}
stuatt$first_9th_school_year_reported <- NULL
idx <- c(2, 8552, 12506) # Specify which SIDs are interesting
stuatt[stuatt$sid %in% idx,]
```

#### Step 1 
*Create one consistent value for gender for each student across years*

```{r echo=TRUE}
library(dplyr) # this is the state-of-the-art in data manipulation in R
library(magrittr) # to improve code readability
stuatt %>% arrange(sid, school_year) %>% 
  select(sid, school_year, male) %>% 
  filter(sid %in% idx)

```

1. *Create a variable that shows how many unique values male assumes for each 
student. Name this variable `nvals_male`. Tabulate the variable and browse 
the relevant data.*


```{r}
stuatt <- stuatt %>% group_by(sid) %>% 
  mutate(nvals_male = length(unique(male))) %>% ungroup() 
table(stuatt$nvals_male)

stuatt %>% select(sid, school_year, male, nvals_male) %>% 
  filter(nvals_male > 1) 

# Or in RStudio
stuatt %>% select(sid, school_year, male, nvals_male) %>% 
  filter(nvals_male > 1) %>% View


```


2. *Identify the modal gender. If multiple modes exist for a student, report the 
most recent gender recorded*

```{r}
# First we need to create a 'mode' function in R that mimics Stata
# statamode creates a list of the modal values and assigns "."
# If more than one mode exists
statamode <- function(x) {
  z <- table(as.vector(x))
  m <- names(z)[z == max(z)]
  if(length(m) == 1){
    if(class(x) %in% c("numeric", "integer", "logical")){
      class(m) <- class(x)
    } else {
      class(m) <- "character"
    }
    
    return(m)
  }
  return(NA)
}

stuatt <- stuatt %>% group_by(sid) %>% 
  mutate(nvals_male = length(unique(male)), 
         male_mode = statamode(male)) %>% ungroup() 

stuatt %>% select(sid, male, male_mode, nvals_male) %>% 
  filter(sid %in% idx)

# In R, NA is equivalent to "." in Stata

```

```{r}
# Replace male with male_mode where male_mode is not missing
stuatt$male[!is.na(stuatt$male_mode)] <- stuatt$male_mode[!is.na(stuatt$male_mode)]
# In R we replace by vector so both sides of the <- have to have the same filter 
# so they are the same length

idx <- c(8552, 12506)

stuatt %>% select(sid, school_year, male, nvals_male, male_mode) %>% 
  filter(sid %in% idx)


# If multiple modes exist, report the most recent gender recorded

stuatt %<>% arrange(sid, school_year) %>% 
  group_by(sid) %>% 
  mutate(temp_male_last = male[school_year == max(school_year)])

# Show sid 12506
stuatt %>% select(sid, school_year, male, nvals_male, male_mode, temp_male_last) %>% 
  filter(sid == 12506)

# Assing temp_male_last to the male variable in cases where no mode exists
stuatt$male[is.na(stuatt$male_mode)] <- stuatt$temp_male_last[is.na(stuatt$male_mode)]


stuatt %>% select(sid, school_year, male, nvals_male, male_mode, temp_male_last) %>% 
  filter(sid == 12506)

# Drop temporary variables

stuatt %<>% select(-nvals_male, -male_mode, -temp_male_last)

```

Let's check we got it right

```{r}
table(stuatt$male)

# Check nvals without creating the variable

stuatt %>% ungroup %>% 
  group_by(sid) %>% 
  summarize(nvals = length(unique(male))) %>% select(nvals) %>% 
  table

length(unique(stuatt$sid))

```


##### Step 2
*Create one consistent value for race_ethnicicty for each student across years*

1. Recode the raw `race_ethnicity` variable as a numeric variable and label it. 
Replace the string race_ethnicity variable with the numeric one.


- 1 = African American, not Hispanic
- 2 = Asian American
- 3 = Hispanic
- 4 = American Indian
- 5 = White, not Hispanic
- 6 = Multiple / Other

```{r echo=TRUE}
# When R reads in Stata files using haven it creates a data type called 
# labelled, for compatibility with Stata and most R functions, we convert 
# this into a more standard factor variable

# Create a copy
stuatt$race_num <- stuatt$race_ethnicity
stuatt$race_ethnicity <- as_factor(stuatt$race_ethnicity)

table(stuatt$race_ethnicity) #check current values

stuatt$race_num <- NA
stuatt$race_num[stuatt$race_ethnicity=='B'] <- 1
stuatt$race_num[stuatt$race_ethnicity=='A'] <- 2
stuatt$race_num[stuatt$race_ethnicity=='H'] <- 3
stuatt$race_num[stuatt$race_ethnicity=='NA'] <- 4
stuatt$race_num[stuatt$race_ethnicity=='W'] <- 5
stuatt$race_num[stuatt$race_ethnicity=='M/O'] <- 6
table(stuatt$race_num)

idx <- c(8552)

stuatt %>% filter(sid %in% idx) %>% 
  select(sid, school_year, race_ethnicity, race_num)

```

```{r echo=TRUE}
# In R categorical variables are best represented as factors
# Factors can have values, and labels
# Create a labeled factor for the new race_num variable

stuatt$race_num2 <- factor(stuatt$race_num, 
                           labels = c('Black', 'Asian', 'Hispanic', 
                                    'Native American', 'White', 'MultipleOther'))

# Compare them to check using a cross-tabulation
table(stuatt$race_ethnicity, stuatt$race_num2)
# Replace them
stuatt$race_ethnicity <- stuatt$race_num2
stuatt$race_num2 <- NULL

table(stuatt$race_ethnicity) # counts
prop.table(table(stuatt$race_ethnicity))*100 #percentages
```


Check:  What does the distribution of your `race_ethnicity` variable look like? 
Let's redraw the tables above in a more readable format.

```{r echo=TRUE, results="markup"}
library(pander) # library to beautify output
pander(prop.table(table(stuatt$race_ethnicity))*100, style = "rmarkdown")
pander(table(stuatt$race_ethnicity), style = "rmarkdown")
```


Let's also draw a figure to show this distribution.

```{r echo=TRUE, fig.align='center'}
library(ggplot2) # the best R library for plotting

qplot(stuatt$race_ethnicity,geom='bar') + 
  theme_bw() + labs(x = 'Race/Ethnicity', y = 'Count', 
                    title = "Frequency of Student Race")
```


*2. Create a variable indicating how many unique values `race_ethnicity` 
assumes for each student called `nvals_race`.*

```{r echo=TRUE, eval=TRUE}
stuatt <- stuatt %>% group_by(sid) %>% 
  mutate(nvals_race = length(unique(race_ethnicity)))

table(stuatt$nvals_race)
```

```{r echo=TRUE}
qplot(as.factor(stuatt$nvals_race), geom='bar') + theme_bw() + 
  xlab('Unique Race Codes') + ylab('Count')

```

*3. Create a variable that shows how many unique values `race_ethnicity` 
assumes for each student and `school_year`. Name this variable `nvals_race_yr`. 
Tabulate the variable and browse the relevant data.*

```{r echo=TRUE, eval=TRUE}
stuatt <- stuatt %>% group_by(sid, school_year) %>% 
  mutate(nvals_race_yr = length(unique(race_ethnicity)))

table(stuatt$nvals_race_yr)

stuatt %>% select(sid, school_year, race_ethnicity, nvals_race, nvals_race_yr) %>%
  filter(nvals_race_yr > 1)

```


4. If more than one race is reported in the same school_year, report students as 
multiracial, unless one of their reported race_ethnicity values is Hispanic. 
Report the student as Hispanic in that case.

```{r}
# Generate a temporary hispanic variable
stuatt$temp_ishispanic <- ifelse(stuatt$race_num == 3 & stuatt$nvals_race_yr > 1, 1, 0)

stuatt %>% select(sid, school_year, race_ethnicity, nvals_race, nvals_race_yr, 
                  temp_ishispanic) %>%
  filter(nvals_race_yr > 1)

stuatt %<>% group_by(sid, school_year) %>% 
  mutate(ishispanic = max(temp_ishispanic, na.rm=TRUE))

stuatt %>% select(sid, school_year, race_ethnicity, nvals_race, nvals_race_yr, 
                  temp_ishispanic, ishispanic) %>%
  filter(nvals_race_yr > 1)

# Replace hispanic values

stuatt$race_num[stuatt$nvals_race_yr > 1 & stuatt$ishispanic == 1] <- 3
stuatt$race_ethnicity[stuatt$nvals_race_yr > 1 & stuatt$ishispanic == 1] <- "Hispanic"
stuatt$race_num[stuatt$nvals_race_yr > 1 & stuatt$ishispanic != 1] <- 6
stuatt$race_ethnicity[stuatt$nvals_race_yr > 1 & stuatt$ishispanic != 1] <- "MultipleOther"

# Drop the temporary variables
stuatt <- select(stuatt, -ishispanic, -temp_ishispanic)

# Drop the duplicates resulting from fixing student with different race_ethnicity
# within a school year

stuatt <- bind_rows(stuatt %>% filter(nvals_race_yr < 2), 
                    stuatt %>% filter(nvals_race_yr > 1) %>% 
                      distinct(sid, school_year, race_ethnicity, .keep_all=TRUE))
stuatt <- select(stuatt, -nvals_race_yr)
# Re arrange
stuatt %<>% arrange(sid, school_year)
```

##- TODO -## Check that this worked

*5 Report the modal race. If multiple modes exist for a student, report the most 
recent race recorded.*

```{r echo=TRUE, results='markup'}
# Create new data frame for individual student
# Create nvals while we are at it

stuatt %<>% group_by(sid) %>% 
  mutate(race_mode = statamode(race_ethnicity))

# tab1 <- table(modes$race_temp,modes$nvals)
# addmargins(tab1, FUN=list(Total=sum), quiet=TRUE)

stuatt %>% filter(sid == 8552) %>% 
  select(sid, school_year, race_ethnicity, nvals_race, race_mode)

stuatt$race_ethnicity[!is.na(stuatt$race_mode)] <- stuatt$race_mode[!is.na(stuatt$race_mode)]

stuatt %>% filter(sid == 8552) %>% 
  select(sid, school_year, race_ethnicity, nvals_race, race_mode)


# Consider cases where the mode is not unique
stuatt %>% filter(sid == 2) %>% 
  select(sid, school_year, race_ethnicity, nvals_race, race_mode)

```


```{r}
# If multiple modes exist report th emost recent race recorded

stuatt %<>% group_by(sid) %>% 
  mutate(race_last = race_ethnicity[school_year == max(school_year)])

stuatt %>% filter(sid == 2) %>% 
  select(sid, school_year, race_ethnicity, nvals_race, race_mode, race_last)

stuatt$race_ethnicity[is.na(stuatt$race_mode)] <- stuatt$race_last[is.na(stuatt$race_mode)]

stuatt %>% filter(sid %in% c(8552, 2)) %>% 
  select(sid, school_year, race_ethnicity)

# Drop temporary variables

stuatt %<>% select(-nvals_race, -race_mode, -race_last, -race_num)
```


```{r}
table(stuatt$race_ethnicity)
```

#### Step 3
*Create consistent values for high school diploma variables.*

1. Recode the `hs_diploma_type variable` as a numeric variable and label it.
Replace the string `hs_diploma_type` variable with the numeric one. Use lower
numbers for more competitive diploma types.

```{r}
# Generate a numeric diploma variable
# Not strictly necessary in R

stuatt$dipl_num <- 4
stuatt$dipl_num <- ifelse(stuatt$hs_diploma_type == "College Prep Diploma", 
                          1, stuatt$dipl_num)
stuatt$dipl_num <- ifelse(stuatt$hs_diploma_type == "Standard Diploma", 
                          2, stuatt$dipl_num)
stuatt$dipl_num <- ifelse(stuatt$hs_diploma_type == "Alternative Diploma", 
                          3, stuatt$dipl_num)


stuatt %>% select(sid, school_year, hs_diploma, hs_diploma_date, 
                  hs_diploma_type, dipl_num) %>% 
  filter(sid == 16)

stuatt$hs_diploma_type <- NULL
stuatt$hs_diploma_type <- stuatt$dipl_num
stuatt$dipl_num <- NULL

stuatt %>% select(sid, school_year, hs_diploma, hs_diploma_date, 
                  hs_diploma_type) %>% 
  filter(sid == 16)

# Now identify the first diploma date reported

stuatt %<>% arrange(sid, hs_diploma_date)

stuatt %<>% group_by(sid) %>% 
  mutate(earliest_diploma_date = min(hs_diploma_date, na.rm=TRUE))


stuatt %>% select(sid, school_year, hs_diploma, hs_diploma_date, 
                  hs_diploma_type, earliest_diploma_date) %>% 
  filter(sid == 16)

#3 Create a variable that shows the earliest diploma type
## This won't work because there may be ties
# stuatt %<>% group_by(sid) %>% 
#  mutate(earliest_dipl_type = hs_diploma_type[hs_diploma_date==earliest_diploma_date])

stuatt %<>% group_by(sid) %>%
 mutate(earliest_dipl_type_mode = statamode(hs_diploma_type[hs_diploma_date==earliest_diploma_date]))


stuatt %>% select(sid, school_year, hs_diploma, hs_diploma_date, 
                  hs_diploma_type, earliest_diploma_date, earliest_dipl_type_mode) %>% filter(sid == 16)


#4 Create a variable that shows the number of unique diploma types recorded
# for the first diploma date

stuatt %<>% group_by(sid) %>%
 mutate(nvals_dipl_type = 
          length(unique(hs_diploma_type[hs_diploma_date==earliest_diploma_date])))


stuatt %>% select(sid, school_year, hs_diploma_type, earliest_diploma_date, 
                  earliest_dipl_type_mode, nvals_dipl_type) %>% 
  filter(sid %in% c(16, 20, 80))


# 5. Identify the modal diploma type. If multiple modes exist for a
# student, report the diploma type in the earliest school year for 
# the first diploma date

stuatt %<>% group_by(sid) %>% 
  mutate(earliest_dipl_type_syear = hs_diploma_type[school_year == min(school_year)])


stuatt %>% select(sid, school_year, hs_diploma_type, earliest_diploma_date, 
                  earliest_dipl_type_mode, nvals_dipl_type, 
                  earliest_dipl_type_syear) %>% 
  filter(sid %in% c(16, 20, 80))


stuatt %<>% group_by(sid) %>%
  mutate(earliest_dipl_type_syear_mode = statamode(earliest_dipl_type_syear))

stuatt %>% 
  select(sid, school_year, hs_diploma_type, earliest_diploma_date, 
                  earliest_dipl_type_mode, nvals_dipl_type, 
                  earliest_dipl_type_syear, 
         earliest_dipl_type_syear_mode) %>% 
  filter(sid %in% c(16, 20, 80))

# 6. If multiple diploma types were recorded for the same school year and first 
# diploma date, report the most competitive diploma type


stuatt %<>% group_by(sid) %>%
  mutate(temp_most_compet = min(earliest_dipl_type_syear))

stuatt %>% 
  select(sid, school_year, hs_diploma_type, earliest_diploma_date, 
                  earliest_dipl_type_mode, nvals_dipl_type, 
                  earliest_dipl_type_syear,  
         earliest_dipl_type_syear_mode, temp_most_compet) %>% 
  filter(sid %in% c(16, 20, 80)) %>% as.data.frame()


# Replace original diploma type variable starting with most specific case, and 
# working backward

stuatt$nvals_dipl_type <- NULL

stuatt$hs_diploma_type[!is.na(stuatt$temp_most_compet)] <- 
  stuatt$temp_most_compet[!is.na(stuatt$temp_most_compet)]

stuatt$hs_diploma_type[!is.na(stuatt$earliest_dipl_type_syear_mode)] <- 
  stuatt$earliest_dipl_type_syear_mode[!is.na(stuatt$earliest_dipl_type_syear_mode)]

stuatt$hs_diploma_type[!is.na(stuatt$earliest_dipl_type_mode)] <- 
  stuatt$earliest_dipl_type_mode[!is.na(stuatt$earliest_dipl_type_mode)]

stuatt %>% 
    select(sid, school_year, hs_diploma_type, earliest_diploma_date, 
           earliest_dipl_type_mode, 
           earliest_dipl_type_syear,  
           earliest_dipl_type_syear_mode, temp_most_compet) %>% 
    filter(sid %in% c(16, 20, 80)) %>% as.data.frame()

# 7. If there are any missing diploma types, mark these as an unknown 
# diploma type

stuatt$hs_diploma_type[is.na(stuatt$hs_diploma_type) &
                         !is.na(stuatt$hs_dipoma_date)] <- 4

# 8. Finally, replace hs_diploma_date with the first hs_diploma_date

stuatt$hs_diploma_date <- stuatt$earliest_diploma_date

# 9. Make sure that diploma is set to 1 if there is a diploma date reported

stuatt$hs_diploma[!is.na(stuatt$hs_diploma_date)] <- 1

# 10. Drop all temporary variables we created

stuatt %<>% select(-earliest_diploma_date, -earliest_dipl_type_mode, 
                   -earliest_dipl_type_syear, -earliest_dipl_type_syear_mode, 
                   -temp_most_compet)


```


Step 4 Drop any unneeded variables, drop duplicates, check the data, and save 
the file

```{r}
# Drop school year as you no longer need it
stuatt %<>% select(-school_year, -birth_date)

# Drop duplicate values

tmp <- stuatt[!duplicated(stuatt),]

# Check that the file is unique by sid

nrow(tmp) == length(unique(stuatt$sid))

# Deduplicate

rm(tmp)
stuatt <- stuatt[!duplicated(stuatt),]

# Save the current file as Student_Attributes.rda

# save(stuatt, file = "Student_Attributes.rda")

# Clean up the workspace
rm(con, idx, tmpfileName, stuatt)
```


## Task 2: STUDENT SCHOOL YEAR
### PURPOSE

In Task 2: Student School Year, you will take the `Student_Classifications_Raw` 
file and generate a clean `Student_School_Year` output file that matches the
specification in Identify with one observation per student and school year. To 
do so, you will first ensure only one grade level is assigned per student per 
school year. Then, you will process the free or reduced price lunch (FRPL)
variable (a proxy for students’ poverty status), individualized education 
program (IEP) variable, English language learner (ELL) variable, and gifted 
variable. You will also examine the total days enrolled, days absent, and 
days suspended variables. The core of this task:

1. Resolve instances when students have more than one grade level in a school
year
2. Keep the highest value of FR PL reported by student by school year
3. If a student has both “has IE P” and “no IE P” reported in a school year, 
keep "has IEP"
4. If a student has both "has ELL" and "no ELL"" reported in a school year, 
keep "has ELL"
5. If a student is observed as both gifted eligible and not eligible, report 
eligible
6. Explore `days_enrolled`, `days_absent` and `days_suspended`
7. Drop duplicate observations to make the file unique by student and school 
year After this, you will have a data set unique by student and school year 
that allows you to assign students to the appropriate ninth grade cohort in 
Task  3.

### HOW TO START

To begin, open the `Student_Classifications_Raw` file in Stata. If you do 
not have Stata, you can follow the steps of the task by looking at the 
instructions and data snippets we have provided.

If this is your first time attempting Task 2, start with the provided raw 
input file. This file teaches you SDP’s cleaning methodology and allows you to 
check answers from a common dataset.

```{r echo=TRUE}
# Read in Stata
library(haven) # required for .dta files

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_Classifications_Raw.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stuclass <- read_stata(con) # read data in the data subdirectory
glimpse(stuclass)
```

### Step 1 Create One Consistent Grade Level for Each Student Within the Same Year

```{r}
# Step 1: Create one consistent grade level for each student within the same year

# Keep the highest grade_level when a student has multiple grade levels 
# within the same year

# distinct values function
nvals <- function(x){
  length(unique(x))
}


varIdx <- c("sid", "school_year", "grade_level", "nvals_grade", 
            "max_grade_level")

stuclass %<>% group_by(sid, school_year) %>% 
  mutate(nvals_grade = nvals(grade_level), 
         max_grade_level = max(grade_level))


stuclass %>% select(one_of(varIdx)) %>% 
  filter(sid == 3)

stuclass$grade_level[stuclass$nvals_grade > 1] <- stuclass$max_grade_level[stuclass$nvals_grade > 1]

stuclass %>% select(one_of(varIdx)) %>% 
  filter(sid == 3)


stuclass %<>% select(-nvals_grade, -max_grade_level)

```

Step 2: Create one consistent FRPL value for each student in the same student-year

```{r}
# 1. Recode raw frpl variable with string type to numeric type

stuclass$frpl_num <- NA
stuclass$frpl_num[stuclass$frpl == "N"] <- 0
stuclass$frpl_num[stuclass$frpl == "R"] <- 1
stuclass$frpl_num[stuclass$frpl == "F"] <- 2

stuclass %>% select(sid, school_year, grade_level, frpl, frpl_num) %>% 
  filter(sid == 80)


stuclass$frpl <- NULL
stuclass$frpl <- stuclass$frpl_num
stuclass$frpl_num <- NULL

stuclass %>% select(sid, school_year, grade_level, frpl) %>% 
  filter(sid == 80)

# 2. Ensure that frpl is consistent by sid and school_year. In cases where 
#  multiple values exist, report the highest value. Follow the same procedure 
#  as Step 1 for grade_level.

# Check if there are any cases where different values of frpl status are reported
#  in a year

stuclass %<>% group_by(sid, school_year) %>% 
  mutate(nvals_frpl = nvals(frpl))

table(stuclass$nvals_frpl)

# Report the highest value of frpl by year for each student, selecting 
# free over reduced over not participating

stuclass %<>% group_by(sid, school_year) %>% 
  mutate(highest_frpl = max(frpl))

stuclass$frpl <- stuclass$highest_frpl

# Label the values so they are easy to understand

# drop the temporary values we created

stuclass %<>% select(-nvals_frpl, -highest_frpl)

```


** Step 3: Create one consistent IEP value for each student within the same year. **

```{r}
# Follow the same procedure as Step 1 for grade_level.
# Report the highest value of iep by year for each student, 
#  selecting has iep over not iep.

stuclass %<>% group_by(sid, school_year) %>% 
  mutate(highest_iep = max(iep)) %>% 
  ungroup() %>% 
  mutate(iep = highest_iep) %>% 
  select(-highest_iep)
```

```{r}
# /*** Step 4: Create one consistent ELL value for each student within the same year. ***/
# Follow the same procedure as Step 1 for grade_level.
# // Report the highest value of ell by year for each student, selecting is ell over not ell.
stuclass %<>% group_by(sid, school_year) %>% 
  mutate(highest_ell = max(ell)) %>% 
  ungroup() %>% 
  mutate(ell = highest_ell) %>% 
  select(-highest_ell)
```


```{r}
# /*** Step 5: Create one consistent gifted value for each student within the same year. ***/
# Follow the same procedure as Step 1 for grade_level.
# // Report the highest value of gifted by year for each student, selecting is enrolled in gifted program over not enrolled.

stuclass %<>% group_by(sid, school_year) %>% 
  mutate(highest_gifted = max(gifted)) %>% 
  ungroup() %>% 
  mutate(gifted = highest_gifted) %>% 
  select(-highest_gifted)

```

```{r}
# /**** Step 6: Drop any unneeded variables, drop duplicates, and save the file ****/
# // 1. Drop duplicate observations

stuclass <- stuclass[!duplicated(stuclass),]

# // 2. Make sure your file is now unique by student and school year

nrow(stuclass) == nvals(paste0(stuclass$sid, stuclass$school_year))

# // 3. Save the current file as Student_School_Year.dta which you will need for Task 3.

# save(stuclass, file = "Student_School_Year.rda")

# Clean up the workspace
rm(con, tmpfileName, stuclass, varIdx)
```


## Task 3: IDENTIFYING THE NINTH-GRADE COHORT
### PURPOSE

In Task 3: Identifying the Ninth Grade Cohort, you will identify the school 
year students first appear in ninth grade using the clean 
`Student_School_Year` research file from Task 2. This essential step allows 
you to form student cohorts and examine longitudinal college-going outcomes.

The core of this task:
1. Flag the first school year a student enrolls in grades 9, 10, 11, or 12.
2. Identify the school year in which the student was first observed in 9th
grade.
3. Impute the school year in which transfer students would have been in grade 9.
4. Replace the first_9th_school_year_observed with the correctly imputed values.
After completing this task, you will have a clean Student_School_Year file that
identifies first-time ninth graders. This file is used both to assemble the
analysis file in Connect and to complete Task 4.

### HOW TO START

To begin, open the `Student_School_Year` file, just created in Task 2, in Stata.
If you do not have Stata, you can follow the steps of the task by looking at 
the instructions and data snippets we have provided.If this is your first 
time attempting Task 3, start with the cleaned output file from Task 2. This 
file teaches you SDP’s cleaning methodology and allows you to check 
answers from a common dataset.

### DATA DESCRIPTION

The input file in this case, `Student_School_Year`, also the output from Task 2,
now follows the structure of `Student_School_Year` in Identify so it is 
unique by `sid` and `school_year`. The aim of this task will be to create a 
`first_9th_school_year_observed` variable using the variables in the file.

Uniqueness: This dataset was cleaned in Task 2 and is now unique by `sid` 
and `school_year`.

```{r}
# Read in Stata
library(haven) # required for .dta f;iles

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "clean/Student_School_Year.dta"
con <- unz(description = "data/clean.zip", filename = tmpfileName, 
           open = "rb")
stusy <- read_stata(con) # read data in the data subdirectory
glimpse(stusy)
```


## - TODO - ## 
Check on p. 35-36 of the guide, incosistency on whether observed_g indicator 
is unique by student, or by student-grade. I've coded it unique by student.

```{r}
# /*** Step 1: Flag the first school year a student enrolls in grades 9, 10, 11, or 12. ***/
# // Create four binary indicators to flag the first school year a student enrolls in grades 9, 10, 11, or 12.

stusy %>% filter(sid == 1) %>% select(sid, school_year, grade_level)

stusy %<>% group_by(sid, grade_level) %>% 
  mutate(tmpG = ifelse(school_year == min(school_year), 1, NA),
         observed_g = 1)

stusy %>% filter(sid == 1) %>% select(sid, school_year, grade_level, 
                                      tmpG, observed_g)
# Use tidyr to spread first_flag and observed_g out
library(tidyr)
stusy$first_flag <- stusy$grade_level


stusy <- spread(stusy, key = first_flag, value = tmpG, sep = "") %>% 
  select(-one_of("first_flag3", "first_flag5", "first_flag6", "first_flag7", 
                 "first_flag8", "first_flag13", "first_flag17"))

stusy$observed <- stusy$grade_level

# Fill in a 1 because we want the observed vectors to populate all values 
# for a student
stusy <- spread(stusy, key = observed, value = observed_g, sep = "_") %>% 
  select(-one_of("observed_3", "observed_5", "observed_6", 
                 "observed_7", "observed_8", "observed_13", 
                 "observed_17")) %>% 
  group_by(sid) %>% 
  mutate(observed_9 = max(observed_9, na.rm=TRUE), 
         observed_10 = max(observed_10, na.rm=TRUE), 
         observed_11 = max(observed_11, na.rm=TRUE), 
         observed_12 = max(observed_12, na.rm=TRUE)) %>% 
  mutate(observed_9 = ifelse(is.finite(observed_9), 1, 0),
         observed_10 = ifelse(is.finite(observed_10), 1, 0),
         observed_11 = ifelse(is.finite(observed_11), 1, 0),
         observed_12 = ifelse(is.finite(observed_12), 1, 0))

# Check how many students are identified as enrolled in grades 9, 10, 11, or 12

tmp <- stusy %>% 
  select(num_range(prefix= "observed_", range = 9:12)) %>% 
  distinct(.keep_all=TRUE)

table(tmp$observed_9)
table(tmp$observed_10)
table(tmp$observed_11)
table(tmp$observed_12)
rm(tmp)

```

** Step 2: Identify the school year in which the student was first observed in 9th grade. **

```{r}
# // Create a variable that lists the first school year a student is observed as enrolled in grade 9.

stusy %<>% group_by(sid) %>% 
  mutate(first_9th_schyear_obs = min(school_year[grade_level == 9]))

# work around weird way R handles minimum of an empty vector

stusy$first_9th_schyear_obs[!is.finite(stusy$first_9th_schyear_obs)] <- NA

stusy %>% filter(sid == 1) %>% 
  select(sid, school_year, grade_level, first_flag9, observed_9, first_9th_schyear_obs)

stusy %>% ungroup %>% distinct(sid, first_9th_schyear_obs) %>% 
  select(first_9th_schyear_obs) %>% unlist %>% table


# Say something about missing values in the list...

```

** Step 3: Impute the school year in which transfer students would have been in grade 9. **

```{r}
# // Impute first_9th_school_year_observed as school_year - 1, school_year - 2, or school_year - 3 for students first observed in 10th, 11th or 12th grade as transfer-ins

stusy$first_flag10[!is.finite(stusy$first_flag10)] <- 0
stusy$first_flag11[!is.finite(stusy$first_flag11)] <- 0
stusy$first_flag12[!is.finite(stusy$first_flag12)] <- 0

stusy$tempfirst9year <- ifelse(stusy$first_flag10 == 1, 
                               stusy$school_year - 1, 
                               ifelse(stusy$first_flag11 == 1, 
                               stusy$school_year - 2, 
                               ifelse(stusy$first_flag12 == 1, 
                               stusy$school_year - 3, 
                               NA)))

stusy %>% filter(sid == 2) %>% 
  select(sid, school_year, grade_level, first_9th_schyear_obs, 
         tempfirst9year)

# What is up with 2003 in the table here in Stata documentation

stusy %<>% group_by(sid) %>% 
  mutate(tempfirst9year = min(tempfirst9year, na.rm=TRUE))

stusy$first_9th_schyear_obs[is.na(stusy$first_9th_schyear_obs) & !is.na(stusy$tempfirst9year)] <- stusy$tempfirst9year[is.na(stusy$first_9th_schyear_obs) & !is.na(stusy$tempfirst9year)]

stusy$tempfirst9year <- NULL

# 
# stusy <- bind_rows(stusy %>% filter(observed_9 == 0), 
#                    stusy %<>% filter(observed_9 == 1) %>%
#   group_by(sid) %>%
#   mutate(first_9th_schyear_obs = max(first_9th_schyear_obs))
# )
# stusy %<>% arrange(sid, school_year)

 
# // Review the distribution of first_9th_school_year_observed for students who transferred in grades 10-12

stusy %>% ungroup %>% 
  filter(first_flag10 > 0) %>% 
  filter(observed_9 == 0) %>% 
  distinct(sid, first_9th_schyear_obs) %>% 
  select(first_9th_schyear_obs) %>% unlist %>% table


stusy %>% ungroup %>% 
  filter(first_flag11 ==1) %>% 
  filter(observed_9 == 0 & observed_10 == 0 & observed_11 == 1) %>% 
  distinct(sid, first_9th_schyear_obs) %>% 
  select(first_9th_schyear_obs) %>% unlist %>% table


stusy %>% ungroup %>% 
  filter(first_flag12 ==1) %>% 
  filter(observed_9 == 0 & observed_10 == 0 & observed_11 == 0 & 
           observed_12 == 1) %>% 
  distinct(sid, first_9th_schyear_obs) %>% 
  select(first_9th_schyear_obs) %>% unlist %>% table

# Consider max of this?
stusy %>% select(sid, school_year, grade_level, 
                first_9th_schyear_obs, observed_9) %>% 
 filter(sid == 3)



# Should this be 2005 or 2007?


```

** Step 4: Adjust the imputation of first_9th_school_year_observed for students who appear in a lower grade in a later school year. **

```{r}
# // 1. Flag students who are observed to be in a lower grade in a subsequent school year.

stusy %<>% arrange(sid, school_year) %>% 
  group_by(sid) %>% 
  mutate(grade_lag = lag(grade_level, order_by = school_year)) %>% 
  mutate(grade_flag = ifelse(grade_lag > grade_level & !is.na(grade_lag > grade_level), 1, 0)) %>% 
  mutate(grade_flag_max = max(grade_flag, na.rm=TRUE)) %>% 
  select(-grade_lag)


stusy %>% select(sid, school_year, grade_level, 
                 first_9th_schyear_obs, grade_flag, grade_flag_max) %>% 
  filter(sid == 3) 

# // 2. Flag the first school year in which students appear in high school grades

stusy %<>% group_by(sid) %>% 
  mutate(first_9th_flag = ifelse(school_year == min(school_year[grade_level %in% c(9:12)]), 1, 0))


stusy %>% select(sid, school_year, grade_level, 
                 first_9th_schyear_obs, grade_flag, grade_flag_max, 
                 first_9th_flag) %>% 
  filter(sid == 3) 


# // 4. Replace the first_9th_school_year_observed with the correctly imputed values.

# Need to drop NAs
stusy$temp4_first9year <- NA
stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 10] <- stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 10] - 1

stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 11] <- stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 11] - 2

stusy$temp4_first9year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 12] <- stusy$school_year[stusy$grade_flag_max == 1 & 
                         stusy$first_9th_flag == 1 & 
                         stusy$grade_level == 12] - 3

stusy %<>% group_by(sid) %>% 
  mutate(temp5_first9year = min(temp4_first9year, na.rm=TRUE))


stusy %>% select(sid, school_year, grade_level, 
                 first_9th_schyear_obs, grade_flag, grade_flag_max, 
                 first_9th_flag, temp4_first9year, temp5_first9year) %>% 
  filter(sid == 3) 


stusy$first_9th_schyear_obs[stusy$grade_flag_max == 1 & 
                              !is.na(stusy$temp5_first9year)] <- stusy$temp5_first9year[stusy$grade_flag_max == 1 & 
                              !is.na(stusy$temp5_first9year)]


stusy %>% ungroup %>% distinct(sid, first_9th_schyear_obs) %>% 
  select(first_9th_schyear_obs) %>% unlist %>% table


```

##- TODO Task 3 Slightly Different Numbers than SDP ##-

** Step 5: Keep only variables relevant to future analyses, and save the file. **


```{r}
# // Keep relevant variables

stusy %<>% select(sid, school_year, grade_level, frpl, iep, ell, gifted, 
                  total_days_enrolled, total_days_absent, 
                  days_suspended_out_of_school, first_9th_schyear_obs)

# // Save the current file as Student_School_Year_Ninth.dta.
# save “${clean}\Student_School_Year_Ninth.dta”, replace
rm(stusy)
```


## Task 4: STUDENT SCHOOL ENROLLMENT
### PURPOSE

In Task 4: Student School Enrollment, you will take the Student_School_Enrollment_Raw file and generate the Student_School_Enrollment file that matches the
specification in Identify. After matching Identify, you will take your dataset a few steps further by consolidating overlapping enrollment spells and determining the
last withdrawal code for each student to yield the file Student_School_Enrollment_Clean.
The core of this task:
1. Create a school_start and school_end variable.
2. Remove abnormal enrollment observations with missing enrollment and withdrawal dates along with enrollment or withdrawal dates that are not in the right
order.
3. Consolidate overlapping enrollments by student by school.
4. Update days_enrolled based on the consolidated enrollments using the new enrollment and withdrawal dates.
5. Determine the last withdrawal code for each student. You will use this data in later analyses to determine a student’s end of high school outcomes.
After completing this, you will have a clean Student_School_Enrollment file. This process sets up our analyses for high school graduation and college enrollment
and persistence outcomes.

### HOW TO START

To begin, open the Student_School_Enrollment_Raw file in Stata. If you do not have Stata, you can follow the steps of the task by looking at the instructions and
data snippets we have provided.
If this is your first time attempting Task 4, start with the provided input file. This file teaches you SDP’s cleaning methodology and allows you to check answers
from a common dataset.

```{r}
# Read in Stata
library(haven) # required for .dta f;iles

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_School_Enrollment_Raw.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stuenr <- read_stata(con) # read data in the data subdirectory
glimpse(stuenr)
```


```{r}
# /*** Step 1: Create a school_start and school_end variable ***/
# // In this example, school start is August 1, and school end is July 31 of each school year. This may be different in your agency.

library(lubridate) # handle dates and times in R correctly

stuenr$school_start <- mdy(paste0("08", "01", stuenr$school_year-1))
stuenr$school_end <- mdy(paste0("07", "31", stuenr$school_year))

# In R we have to create a character string that we convert to a date
# Converting numerics to dates and times can introduce errors

# /*** Step 2: Remove abnormal enrollment observations. ***/
# // 1. Drop observations missing both enrollment and withdrawal dates.

stuenr %<>% filter(!is.na(enrollment_date) & !is.na(withdrawal_date))

# // 2. Drop observations with enrollment and withdrawal dates on same day.

stuenr %<>% filter(!enrollment_date == withdrawal_date & !is.na(enrollment_date))

# // 3. Drop observations with withdrawal date earlier than enrollment date.

stuenr %<>% filter(!is.na(withdrawal_date) & !withdrawal_date < enrollment_date)

# // 4. Drop observations with enrollment date after the end of the current school year.

stuenr %<>% filter(school_end > enrollment_date)

# // 5. Drop observations with enrollment date before the beginning of the current school year.

stuenr %<>% filter(school_start <= enrollment_date)

# // 6. Drop observations with withdrawal date more than one month after the end of the school year.

stuenr %<>% filter(withdrawal_date <= (school_end + 31) & !is.na(withdrawal_date))

# // 7. Check to make sure enrollment dates are in the correct school year.

table(stuenr$enrollment_date >= stuenr$school_start)
table(stuenr$enrollment_date <= stuenr$school_end)

```

/*** Step 3: Consolidate overlapping enrollments by student by school. ***/

```{r}
# // 1. Sort enrollment spells in ascending order and then check how many overlapping enrollment spells exist for a student at the same school.


stuenr %<>% arrange(sid, school_code, enrollment_date)


stuenr %<>% group_by(sid, school_code) %>% 
  mutate(lag_withdrawal_date = lag(withdrawal_date)) %>% ungroup %>% 
  group_by(sid, school_code, school_year) %>% 
    mutate(min_enroll_date = min(enrollment_date))

# table(stuenr$enrollment_date <= stuenr$lag_withdrawal_date & 
        # !is.na(stuenr$lag_withdrawal_date))
# 682?

tmp <- stuenr %>% filter(sid == 2) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date, lag_withdrawal_date,
         withdrawal_code_desc, min_enroll_date)

# // 2. For overlapping observations, replace the enrollment date and enrollment code description of all but the first enrollment spell with the earliest enrollment date

stuenr$enrollment_date[stuenr$enrollment_date <= stuenr$lag_withdrawal_date & 
                         !is.na(stuenr$lag_withdrawal_date)] <- stuenr$min_enroll_date[stuenr$enrollment_date <= stuenr$lag_withdrawal_date & 
                         !is.na(stuenr$lag_withdrawal_date)]

stuenr %>% filter(sid == 2) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date,
         withdrawal_code_desc)
```

**// 3. Replace the withdrawal date and withdrawal code description of the earliest enrollment spell with the latest withdrawal date.**


```{r}
# // Sort the data first so that latest withdrawal
# information appears as the first record.

stuenr %<>% arrange(sid, school_code, enrollment_date, withdrawal_date)

stuenr %>% filter(sid == 2) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date,
         withdrawal_code_desc)

# // Replace withdrawal_date
# // Replace withdrawal_code_description

stuenr %<>% group_by(sid, school_code, enrollment_date) %>%
  mutate(withdrawal_date = last(withdrawal_date), 
         withdrawal_code_desc = last(withdrawal_code_desc))

```

** Step 4: Update days_enrolled based on the consolidated enrollments using the new enrollment and withdrawal dates. **

```{r}

stuenr$days_enrolled <- stuenr$withdrawal_date - stuenr$enrollment_date

stuenr %>% filter(sid == 2) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date,
         withdrawal_code_desc, days_enrolled)


```


** Step 5: Determine the last withdrawal code for each student. You will use this data in later analyses to determine a student's end of high school outcomes. **

```{r}

stuenr %<>% arrange(sid, withdrawal_date)

stuenr %>% filter(sid == 16) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date,
         withdrawal_code_desc)

# p.46 withdrawal date wrong in table, or needs to be updated

stuenr %<>% group_by(sid) %>% 
  mutate(last_withdrawal_reason = last(withdrawal_code_desc))

stuenr %>% filter(sid == 16) %>% 
  select(sid, school_year, school_code, enrollment_date, 
         enrollment_code_desc, withdrawal_date,
         withdrawal_code_desc, last_withdrawal_reason)

```


** Step 6: Drop any unneeded variables, drop duplicates, and save the file **

```{r}
# // 1. Drop duplicate records

stuenr %<>% select(-min_enroll_date, -lag_withdrawal_date)

stuenr <- ungroup(stuenr) %>% distinct(sid, school_year, school_code, 
                                    enrollment_date, .keep_all = TRUE)

# // 2. Confirm that file is unique by student, school_year, school_code, and enrollment_date

nvals(paste0(stuenr$sid, stuenr$school_year, stuenr$school_code, 
             stuenr$enrollment_date)) == nrow(stuenr)


# // 3. Save the current file as Student_School_Enrollment_Clean
# save “${clean}\Student_School_Enrollment_Clean.dta”, replace
# save(stuenr, file = "clean/Student_School_Enrollment_Clean.rda")
rm(tmp, stuenr); gc()
```


## Task 4: STUDENT TEST SCORES

### PURPOSE

structure of Identify. Through this task, you will generate three different clean output files that contain a single score and test-taking instance for
each student:
• Prior Achievement (one 8th grade state test score per student),
• SAT scores (one SAT score per student), and
• ACT scores (one ACT score per student).
The file for Prior Achievement will contain students’ achievement on state standardized Math and English Language Arts tests in 8th grade. This
will allow you to control for prior academic achievement when you examine college-going outcomes. The SAT and ACT score files will be used for
defining highly qualified high school graduates.
The core of this task:
• Prior Achievement
1. Clean state test scores and resolve instances where students took the same test multiple times.
2. Standardize test scores to a mean of 0 and a standard deviation of 1. This allows you to compare across tests and years when different score scales were
used.
3. Generate a composite math and English score for scaled and standardized test scores in eighth grade.
• SAT
1. Clean SAT test scores and resolve instances where students took the same test multiple times.
2. Generate a total SAT score based on math, verbal, and writing scores.
• ACT
1. Clean ACT test scores and resolve instances where students took the same test multiple times.
After completing this, you will have a Prior_Achievement file with 8th grade test scores. You will also have SAT and ACT files. All three files will be used in
Connect.


### HOW TO START

the steps of the task by looking at the instructions and data snippets we have provided.
If this is your first time attempting Task 5, start with the provided input file. This file teaches you SDP’s cleaning methodology and allows you to check answers
from a common dataset


### DATA DESCRIPTION

The input file, Student_Test_Scores, follows the structure of Student_Test_Scores in Identify so it is unique by sid, test_code, and test_date. The aim of this task
will be to create three separate clean output files, Prior_Achievement, SAT, and ACT, that report only one test score per student. This means that for eight grade
prior achievement duplicates of the same test taken in the same and different years will need to be resolved. Also any duplicates of SAT or ACT scores will need to
be resolved as well.


Prior Achievement (8th grade state test scores),
Ideally, state test data in its raw form is unique by sid, test_subject, grade_level, and school_year. However, some students re-take the same test for the same
grade in the same year. To fix this, you will make the 8th grade test score data in Student_Test_Scores unique by sid, test_subject, grade_level, and school_
year by removing any same year repeat test taking instances. Then, you will manipulate the data so tests for different subjects in the same grade_level fall on
the same row so the data is unique by sid, test_subject, and grade_level. As a final step, if a student took the same test in different years (e.g. by repeating a
grade), you will take the earliest instance. The data will finally be unique by sid and is considered a clean file and ready to be incorporated into the analysis file in
Connect.

SAT
Ideally, SAT test data in its raw form is unique by sid. However, some students re-take the SAT. To fix this, you will take the data unique by sid, test_subject, and
test_date and reshape it so the data will finally be unique by sid and is considered a clean file and ready to be incorporated into the analysis file in Connect.

ACT
Ideally, ACT test data in its raw form is unique by sid. However, some students re-take the ACT. To fix this, you will take the data unique by sid, test_subject, and
test_date and reshape it so the data will finally be unique by sid and is considered a clean file and ready to be incorporated into the analysis file in Connect.

** Part I: Clean Prior Achievement Scores **

```{r readstutest}
# Read in Stata
library(haven) # required for .dta f;iles


# // 0. Load the Student_Test_Scores data file.
# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_Test_Scores.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stutest <- read_stata(con) # read data in the data subdirectory
glimpse(stutest)

# Convert to R style
stutest$test_subject <- as_factor(stutest$test_subject)
stutest$test_subject <- tolower(as.character(stutest$test_subject))
```

```{r}
# // 1. Keep only the variables you need and limit the sample to state test scores in 8th grade.


statetest <- stutest %>% 
            select(sid, test_type, test_subject, school_year, 
                   grade_level, scaled_score, raw_score, test_date) %>% 
    filter(test_type == "State" & grade_level == 8)

# Leave the original stutest, we will come back to this later. In R 
# we can keep multiple datasets open in the workset at the same time.

# // 2. Clean up raw and scaled scores.
# // Change raw and scaled scores to missing if zero.

for(var in c("raw_score", "scaled_score")){
  statetest[, var][statetest[,var] == 0] <- NA
}

# //Drop observations missing both a raw and scaled test score.

statetest %<>% filter(!is.na(raw_score) | !is.na(scaled_score))

# 3. Identify same-year repeat test takers and take the highest test score.
# For ties in scores, take the last date tested
statetest %<>% arrange(sid, test_subject, 
                       grade_level, school_year, scaled_score)

statetest %>% filter(sid == 595) %>% 
  select(sid, test_type, school_year, test_date, grade_level, 
         test_subject, scaled_score, raw_score)


statetest %<>% group_by(sid, test_subject, school_year, grade_level) %>% 
  mutate(keep_flag = scaled_score == max(scaled_score) & 
           test_date == max(test_date)) %>% 
  ungroup %>% 
  filter(keep_flag) %>% 
  select(-keep_flag)

statetest %>% filter(sid == 595) %>% 
  select(sid, test_type, school_year, test_date, grade_level, 
         test_subject, scaled_score, raw_score)

# // Verify that each student has only one state test in a
# subject in a school year.

statetest %>% distinct(sid, test_subject, grade_level, school_year) %>%
  nrow == nrow(statetest)

# // 4. Reshape the data so math and ELA tests appear on the same row.

statetest <- reshape(as.data.frame(statetest), 
               v.names = c("raw_score", "scaled_score"), 
               timevar = c("test_subject"), 
               idvar = c("sid", "test_type", "test_date",
                         "school_year", "grade_level"),
               direction = "wide", 
               sep = "_")

# // 5. Compute standardized test scores with mean 0 and standard deviation 1.

statetest$scaled_math_std <- scale(statetest$scaled_score_math)
statetest$scaled_ela_std <- scale(statetest$scaled_score_ela)


# // 6. Identify different-year repeat test takers and take the earliest test score.
# 
# // First process ELA scores
# preserve
# drop *_math*
# drop if scaled_score_ela==.
# // Keep only the earliest instance in which the student
# took the test
# sort sid grade_level school_year
# drop if sid==sid[_n-1] & grade_level==grade_
# level[_n-1] & school_year>= school_year[_n-1] & scaled_
# score_ela[_n-1]!=.
# bys sid: gen count=_n
# tab count
# drop count
# // Save the ela_scores as a tempfile to be merged on
# tempfile ela_scores
# save `ela_scores’
# restore
# 
# // Next process math scores
# drop *_ela*
# drop if scaled_score_math==.
# // Keep only the earliest instance in which the student took the test
# sort sid grade_level school_year
# drop if sid==sid[_n-1] & grade_ level==grade_level[_n-1] &
# school_year>= school_year[_n-1] & scaled_score_math[_n-1]!=.
# bys sid: gen count=_n
# tab count
# drop count
# // Merge the ela_scores tempfile onto the math scores
# merge 1:1 sid using `ela_scores’, nogen

statetest %<>% group_by(sid) %>% 
  mutate(keep_flag = test_date == min(test_date)) %>% 
  filter(keep_flag) %>% select(-keep_flag)

# Not sure in R we need to process these separately at all!

```


```{r}
# // 7. Verify that each student has only one state test, and drop unneeded variables.

nrow(statetest) == nvals(statetest$sid)

statetest %<>% select(-test_date, -test_type)

# // 8. Generate composite scaled and standardized scores that average ELA and math scores.

statetest$scaled_score_composite <- (statetest$scaled_score_ela + statetest$scaled_score_math) /2

statetest$scaled_score_composite_std <- (statetest$scaled_math_std + statetest$scaled_ela_std) /2

# // 9. Save the current file as Prior_Achievement.dta.

statetest %<>% arrange(sid, school_year, grade_level) %>% 
  select(sid, school_year, grade_level, raw_score_math, raw_score_ela,
         scaled_score_math, scaled_score_ela, scaled_score_composite,
         scaled_math_std, scaled_ela_std, scaled_score_composite_std)

# save “${clean}/Prior_Achievement.dta”, replace
```


** Part II: Clean SAT Scores **

```{r}
# // 1. Keep only the variables and limit the sample to SAT.
sattest <- stutest %>% filter(test_type == "SAT")

sattest %<>% select(sid, test_subject, test_date, scaled_score)
# // 2. Drop duplicate observations and any observations missing test scores.

sattest %<>% distinct()
sattest %<>% filter(!is.na(scaled_score))

# // 3. Reshape the data so that math, ELA, and writing scores appear on one row by student and test date.

sattest <- reshape(as.data.frame(sattest), 
               v.names = c("scaled_score"), 
               timevar = c("test_subject"), 
               idvar = c("sid", "test_date"),
               direction = "wide", 
               sep = "_")

# Rename for convenience
names(sattest) <- c("sid", "sat_test_date", "sat_math_score", 
                    "sat_verbal_score", "sat_writing_score")

sattest %<>% arrange(sid, sat_test_date)
# // 4. Identify repeat test takers and take the earliest test score.

sattest %<>% group_by(sid) %>% 
  mutate(keep_flag = sat_test_date == min(sat_test_date)) %>% 
  filter(keep_flag) %>% select(-keep_flag)

# // Verify that the file is now unique by student.

nrow(sattest) == nvals(sattest$sid)

# // 5. Verify that test scores from the component subjects are not missing and generate total scores.

table(!is.na(sattest$sat_math_score) & !is.na(sattest$sat_verbal_score))
sattest$sat_total_score <- sattest$sat_math_score + sattest$sat_verbal_score

table(!is.na(sattest$sat_math_score) & !is.na(sattest$sat_verbal_score) & 
        !is.na(sattest$sat_writing_score))

sattest$sat_total_score_plus_writing <- sattest$sat_math_score +
  sattest$sat_verbal_score + sattest$sat_writing_score

# // 6. Save the current file as SAT.dta.

```

** Part III: Clean ACT Scores **


```{r}
# // 1. Keep only the variables you need and limit the sample to ACT.

acttest <- stutest %>% filter(test_type == "ACT")
acttest %<>% select(sid, test_subject, test_date, scaled_score)

# // 2. Identify repeat test takers and take the earliest test score.

acttest %<>% group_by(sid) %>% 
  mutate(keep_flag = test_date == min(test_date)) %>% 
  filter(keep_flag) %>% select(-keep_flag)

# // 3. Keep and rename the relevant variables.
acttest %>% select(sid, test_date, scaled_score)
names(acttest) <- c("sid", "act_test_date", "act_composite_score")

# // Verify that the file is now unique by student.
nrow(acttest) == nvals(acttest$sid)

# // 4. Save the current file as ACT.dta.
rm(stutest, acttest, sattest, statetest)
```

## Task 6: STUDENT CLASS ENROLLMENT

### PURPOSE

In Task 6: Student Class Enrollment, you will take the Class_Raw file and the Student_Class_Enrollment file to create the Student_Class_Enrollment_
Merged file that combines these two files together. The combined file will identify a unique observation by student and class id. To obtain this file,
you will first clean the Class_Raw file to identify core courses in math and ELA based on the course description variable and match the specification
in Identify. This will make the class file unique by class id. Second, you will merge the Class file and the Student Class Enrollment file and make it
unique by student id and class id.
The core of this task:
1. Using the Class file:
a. Drop incomplete observations
b. Flag core math and English courses based on the course description
2. Merging the Student Class Enrollment file:
a. Merge the Class file onto the Student_Class_Enrollment_Raw file
b. Evaluate course marks and drop courses with no record of completion
c. Evaluate course enrollment so that each student has only one enrollment record for a course
The Student_Class_Enrollment_Merged file will be used in Connect to create on-track indicators for students. On-track indicators explore year-byyear
academic progress towards high school graduation and college readiness. For instance, using course credit and course grade information, one
might ask what percent of students earn the minimum number of credits in their core courses to satisfy agency graduation requirements?

### HOW TO START 

To begin, open the Class_Raw file in Stata. This file contains data linking students to teachers. If you do not have Stata, you can follow the steps of the task by
looking at the instructions and data snippets we have provided. In the second part of this task, you will then use the Student_Class_Enrollment file.
If this is your first time attempting Task 6, start with the provided input file. This file teaches you SDP’s cleaning methodology and allows you to check answers
from a common dataset.

### DATA DESCRIPTION FOR RAW FILE 

The input file, Class_Raw, varies from Class in Identify in a number of key ways. Most importantly, the data is not unique by cid as shown in Identify. For instance,
there may be more than one course description that describes the same course. Also, a tid is not included as it is not required for the questions later asked in this
toolkit. Support for a Class file with tid will come with the Human Capital version of the toolkit. The aim of this task then is to eliminate any duplicate course code
descriptions and match the Class file in Identify in its structure and uniqueness so it is unique by cid alone.

```{r readstuclass}
# Read in Stata
library(haven) # required for .dta f;iles

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Class_Raw.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
classRaw <- read_stata(con) # read data in the data subdirectory
glimpse(classRaw)

# TODO - File is named incorrectly on p.59
tmpfileName <- "raw/Student_Class_Enrollment.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stuclass <- read_stata(con) # read data in the data subdirectory
glimpse(stuclass)

```

** Part I: Clean the Class file **

```{r}

# // 1. Identify the critical variables that identify a class.

local_ids <- c("cid", "school_year", "school_code", "section_code", 
                     "course_code")

# // 2. Drop the observations where any of the critical variables are missing
classRaw %<>% 
  filter(complete.cases(.[, local_ids]))

```

** Step 1: Flag core math and English courses. **

```{r}
# // Note that agencies may have varying consistency in course
# names and use different criteria to identify a core course
# vs an elective.
# // In some cases, other criteria may have to be applied to
# identify core courses (e.g. the department the course is
# listed in, or length of the course.)
# // We provide a simplified version of the cleaning process
# for the class file: work within your agency to determine the
# best criteria.

# // 1. Tabulate course names
table(classRaw$course_code_desc)

# // 2. Flag math courses based on the tabulation results
# // Generate a flag variable

classRaw$math_flag <- NA

# // Use the regexm function to identify course names that contain common word stems, but slightly different spellings, e.g.
# Algebra I and Algebra-I

# In R the patterns need to not have spaces in the grep command
# The spaces will be matched
# | = OR
# grep does partial matching
# grpl returns TRUE/FALSE, as.numeric converts this to 1/0
classRaw$math_flag <- as.numeric(grepl("GEOM|ALGEBRA|MATH|STAT|CALC|TRIG", 
                                        classRaw$course_code_desc))
# // Check the results of flagging your variables

table(classRaw$course_code_desc, classRaw$math_flag)

```

```{r}
# // 3. Repeat this process for flagging ELA courses

classRaw$ela_flag <- NA

# In R the patterns need to not have spaces in the grep command
# The spaces will be matched
# | = OR
# grep does partial matching
# grpl returns TRUE/FALSE, as.numeric converts this to 1/0
classRaw$ela_flag <- as.numeric(grepl("ENG|ELA", 
                                        classRaw$course_code_desc))
# // Check the results of flagging your variables

table(classRaw$course_code_desc, classRaw$ela_flag)
```

*** Step 3: Drop any unneeded variables, drop duplicates, and save the temporary file **

```{r}

# // 1. Drop the course_code_desc, as it is no longer needed.

classRaw %<>% select(-course_code_desc)
classRaw %<>% distinct()

# // 2. Verify that the data is unique by cid, and also unique by school year, school code, section code and course code.
nrow(classRaw) == nvals(classRaw$cid)

classRaw %>% distinct(school_year, school_code, 
                      section_code, course_code) %>% 
  nrow == nrow(classRaw)

```


** Part II: Clean the Student_Class_Enrollment file. **

** Step 0: Load the Student_Class_Enrollment data file. **

** Step 1: Merge on the temporary Class file you saved earlier to the Student_Class_Enrollment file **

```{r}
# Merging is fun
# keep only files merged from both files
stuclass <- inner_join(stuclass, classRaw, by = "cid")

# /*** Step 2: Evaluate course marks. ***/


table(stuclass$final_grade_mark, stuclass$credits_possible)
table(stuclass$final_grade_mark, stuclass$final_grade_mark_num)
# 
# // Some letter marks (NGPA and P) indicate that they do not count toward GPA, so you may leave the numeric mark as missing.

# ** Step 3: Evaluate course completion. **
# // Drop observations that have no record of course completion

stuclass %<>% filter(!is.na(final_grade_mark) & 
                       !is.na(final_grade_mark_num) & 
                       !is.na(credits_earned))

# ** Step 4: Evaluate course enrollment. **
# // Fix cases where a student has multiple observations for the same course with the same year and marking period (i.e. with overlapping enrollment dates)


# // 1. Remove enrollment and withdrawal dates that are not in the current school year.


library(lubridate) # handle dates and times in R correctly

stuclass$school_start <- mdy(paste0("08", "01", stuclass$school_year-1))
stuclass$school_end <- mdy(paste0("07", "31", stuclass$school_year))

stuclass$class_enrollment_date[stuclass$class_enrollment_date < 
                                 stuclass$school_start | 
                                 stuclass$class_enrollment_date > 
                                 stuclass$school_end] <- NA
stuclass$class_withdrawal_date[stuclass$class_withdrawal_date < 
                                 stuclass$school_start | 
                                 stuclass$class_withdrawal_date > 
                                 stuclass$school_end] <- NA


stuclass %<>% select(-school_start, -school_end)


stuclass %>% filter(sid == 2251 & cid == 78150780) %>% 
  select(sid, cid, school_code, school_year, class_enrollment_date, 
         class_withdrawal_date)

# // 2. Identify the variables that identify a course

local_ids <- c("sid", "cid", "school_year", "marking_period")

# // 3. Populate all enrollments with the earliest enrollment date

# use group_by_ to pass character values to dplyr
stuclass %<>% ungroup %>% 
  group_by(sid, cid, school_year, marking_period) %>% 
  arrange(class_enrollment_date) %>%
  mutate(first_enroll = min(class_enrollment_date, na.rm=TRUE)) 


# There is a bug here in the enrollment date
# If you use group_by_ and mutate without the underscore
# in a pipe, and then calculate max or min of a date, you 
# get the wrong time

stuclass$class_enrollment_date <- stuclass$first_enroll
stuclass %<>% select(-first_enroll)

stuclass %>% ungroup %>% filter(sid == 2251 & cid == 78150780) %>% 
  select(sid, cid, school_code, school_year, class_enrollment_date, 
         class_withdrawal_date)

# // 4. Populate all enrollments with the latest withdrawal date

stuclass %<>% ungroup %>% 
  arrange(sid, cid, school_year, marking_period, class_withdrawal_date) %>%
  group_by(sid, cid, school_year, marking_period) %>% 
  mutate(last_withdraw = max(class_withdrawal_date, na.rm=TRUE))

stuclass %>% ungroup %>% filter(sid == 2251 & cid == 78150780) %>% 
  select(sid, cid, class_enrollment_date, 
         class_withdrawal_date, last_withdraw)

stuclass$class_withdrawal_date <- stuclass$last_withdraw
stuclass$last_withdraw <- NULL

stuclass %>% ungroup %>% filter(sid == 2251 & cid == 78150780) %>% 
  select(sid, cid, school_code, school_year, marking_period, 
         section_code, class_enrollment_date, 
         class_withdrawal_date, class_withdrawal_date)

```

** Step 5: Drop any unneeded variables, drop duplicates, and save the file **

```{r}
# // 1. Drop duplicate values

stuclass %<>% ungroup %>% distinct()

# // 2. Verify that the file is unique by sid and cid

nrow(stuclass) == nvals(paste0(stuclass$sid, stuclass$cid, sep ="_"))

# // 3. Order the variables

stuclass %<>% select(sid, cid, school_year, school_code, course_code,
                     marking_period, section_code, instructional_level,
                     credits_possible, math_flag, ela_flag,
                     class_enrollment_date, class_withdrawal_date,
                     final_grade_mark, final_grade_mark_num, 
                     credits_earned)
# // 4. Sort the data

stuclass %<>% ungroup() %>% 
  arrange(sid, school_year, marking_period, cid)

# // 5. Save the current file as Student_Class_Enrollment_Merged.dta.


```


## Task 7 STUDENT NSC ENROLLMENT

### PURPOSE

In Task 7: Student NSC Enrollment, you will take the Student_NSC_Enrollment file that matches the specification in Identify and produce a Student_
NSC_Enrollment_Indicators file that includes some of the first college enrollment indicators you will need for further analysis.
College enrollment data is obtained from the National Student Clearninghouse (NSC). NSC matches students from a file your agency sends, including
student id, student name, high school from where the student graduated, graduation date, and some other variables. For more information on the
NSC matching process and requirements, visit http://www.studentclearinghouse.org/high_schools/studenttracker
The core of this task:
1. Rename the variables typically returned by NSC
2. Format the date values
3. Standardize the variables that reflect the type of college the student enrolls in
4. Create a college graduation indicator
5. Interpret the college enrollment status
6. Identify the first college the student attended
After this task, you will merge the Student_NSC_Indicators file onto the college-going analysis file from Connect. You will use this file and the high
school graduation variables you will also create in Connect to then to generate further college-going variables, such as variables that indicating if a
student enrolled in college the fall after graduation, enrolled in college a year after graduation, and persisted through subsequent years of college.

### HOW TO START

To begin, open the Student_NSC_Enrollment file in Stata. This file contains data on college enrollment and persistence for students in your agency. If you do not
have Stata, you can follow the steps of the task by looking at the instructions and data snippets we have provided.
If this is your first time attempting Task 7, start with the provided input file. This file teaches you SDP’s cleaning methodology and allows you to check answers
from a common dataset.


### DATA DESCRIPTION

The input file, Student_NSC_Enrollment, follows the structure of Student_NSC_Enrollment in Identify so it is unique by sid, college_code_branch, enrollment_
begin, and enrollment_end. This usually equates to a semester. Though the exact structure of the data you receive from NSC may vary, it will likely look
something like this. The aim of this task then is to become familiar with the NSC data and start building college enrollment outcomes that will be expanded upon in
Connect.

** Step 0: Load the Student_NSC_Enrollment data file. **

```{r}
# Read in Stata
library(haven) # required for .dta files

# To read data from a zip file we create a connection to the path of the 
# zip file
tmpfileName <- "raw/Student_NSC_Enrollment.dta"
con <- unz(description = "data/raw.zip", filename = tmpfileName, 
           open = "rb")
stunsc <- read_stata(con) # read data in the data subdirectory
glimpse(stunsc)
```

```{r}
# // 1. Rename variables to indicate that they are NSC variables.

names(stunsc) <- c("sid", "n_record_found_yn", "n_enrollment_begin", 
                   "n_enrollment_end", "n_college_opeid", 
                   "n_college_name", "college_state","yr2_yr4", 
                   "public_private", "n_enrollment_status", 
                   "graduated", "n_degree_date", 
                   "n_enrl_sequence", "degree_title", "major")

# // 2. Format the date values as dates.
library(lubridate)

stunsc <- as.data.frame(stunsc)
for(i in c("n_enrollment_begin", "n_enrollment_end", 
           "n_degree_date")){
  stunsc[, i] <- lubridate::ymd(as.character(stunsc[, i]))
  }

stunsc %>% filter(sid == 13047) %>% 
  select(sid, n_record_found_yn, n_enrollment_begin, 
         n_enrollment_end, n_college_name, yr2_yr4, 
         public_private, n_enrollment_status, graduated)

### Drop missing
# stunsc %<>% filter(stunsc$college_state != "")


# // 3. Standardize types of college by:
# // 2-year and 4-year college

stunsc$n_college_4yr <- ifelse(stunsc$yr2_yr4 == "4-year", 1, 0)
stunsc$n_college_2yr <- ifelse(stunsc$yr2_yr4 == "2-year" |
                                 stunsc$yr2_yr4 == "Less Than 2 Years", 
                               1, 0)

stunsc$yr2_yr4 <- NULL

# // Public and private college
table(stunsc$public_private)

stunsc$n_college_public <- ifelse(stunsc$public_private == "Public", 1, 0)
stunsc$n_college_private <- ifelse(stunsc$public_private == "Private", 1, 0)
stunsc$public_private <- NULL

# // In-state and out-of-state college
table(stunsc$college_state)

stunsc$n_college_instate <- ifelse(stunsc$college_state == "MA", 1, 0)
stunsc$n_college_outstate <- ifelse(stunsc$college_state != "MA", 1, 0)
stunsc$college_state <- NULL

# // 4. Create a college graduation indicator.
stunsc$n_degree <- ifelse(stunsc$graduated == "Y", 1, 0)
stunsc$graduated <- NULL
# // 5. Interpret enrollment status.

table(stunsc$n_enrollment_status)

stunsc$n_enrl_status <- factor(stunsc$n_enrollment_status, 
                               levels = c("F", "H", "L", "W", 
                                          "A", "D"))
stunsc$n_enrollment_status <- NULL

```


** Step 2: Identify first college attended by type (any, 4-year and 2-year) that didn’t result in a withdrawal. **

```{r}

# // 1. Specify these types of college (any, 4-year, 2-year) in globals.
# global condition_any “n_college_4yr == 1 | n_college_2yr == 1”
# global condition_4yr “n_college_4yr == 1”
# global condition_2yr “n_college_2yr == 1”
# // 2. Calculate the days enrolled.

stunsc$days_enrolled <- stunsc$n_enrollment_end - stunsc$n_enrollment_begin

# // 3. Identify the first college a student enrolled in by type (any, 2-year, and 4-year).

stunsc %>% filter(sid == 13047) %>% 
  select(sid, n_record_found_yn, n_enrollment_begin, 
         n_enrollment_end, n_college_name,  
         n_enrl_status, n_college_4yr, n_college_2yr)

stunsc %<>% group_by(sid) %>% 
  mutate(flag_status = ifelse(n_enrl_status %in% c("F", "H", "L"), 1, 0))

stunsc %>% filter(sid == 13047) %>% 
  select(sid, n_record_found_yn, n_enrollment_begin, 
         n_enrollment_end, n_college_name,  
         n_enrl_status, n_college_4yr, n_college_2yr, flag_status)

stunsc %<>% group_by(sid, n_college_4yr) %>% 
  mutate(first_enr_date_4yr = min(n_enrollment_begin[flag_status > 0])) %>%
  ungroup %>%
  group_by(sid, n_college_2yr) %>%
  mutate(first_enr_date_2yr = min(n_enrollment_begin[flag_status > 0])) %>%
  ungroup %>%
  group_by(sid) %>%
  mutate(first_enr_date_any = min(n_enrollment_begin[flag_status > 0])) 
  

stunsc %>% filter(sid == 13047) %>% 
  select(sid,n_enrollment_begin, 
         n_enrollment_end,   
         n_enrl_status, n_college_4yr, n_college_2yr, flag_status, 
         first_enr_date_2yr, first_enr_date_4yr, 
         first_enr_date_any) %>% as.data.frame


stunsc %>% filter(sid == 13047) %>% 
  select(sid, n_college_opeid,  
         n_enrl_status, n_college_4yr, n_college_2yr, flag_status, 
         first_enr_date_2yr, first_enr_date_4yr, 
         first_enr_date_any) %>% as.data.frame

stunsc %<>% group_by(sid) %>% 
  mutate(first_college_any_opeid = n_college_opeid[first_enr_date_any == n_enrollment_begin][1],
         first_college_4yr_opeid = n_college_opeid[first_enr_date_4yr == n_enrollment_begin & n_college_4yr > 0][1], 
         first_college_2yr_opeid = n_college_opeid[first_enr_date_2yr == n_enrollment_begin & n_college_2yr > 0][1]) %>% 
  ungroup


# # // Get the college name and id for the first enrollment date
# 
# stunsc %>% filter(sid == 13047) %>% 
#   select(sid, n_college_opeid,  
#          n_enrl_status, n_college_4yr, n_college_2yr,  
#          first_enr_date_2yr, first_enr_date_4yr, 
#          first_enr_date_any, 
#          first_college_any_opeid, 
#          first_college_4yr_opeid, first_college_2yr_opeid) %>% as.data.frame
# 
# # // Count how many first college names and ids you got for each student
# # bys sid: egen nvals_first_college_`var’_`type’ = nvals(temp_first_college_`var’_`type’)
# 
# tmp <- stunsc %>% select(sid, first_college_any_opeid, 
#                          first_college_4yr_opeid, first_college_2yr_opeid)
# 
# tmp <- tmp[!duplicated(tmp), ]
# tmp$nunique <- apply(tmp[, 2:4], 1, function(x) length(unique(na.omit(x))))
# 
# stunsc <- left_join(stunsc, tmp[, c(1,5)], "sid")
# rm(tmp); gc()
# 
# 
# stunsc %>% filter(nunique > 1) %>% 
#   select(sid, n_enrl_status, n_college_4yr, n_college_2yr, 
#          n_enrollment_begin,
#          n_college_opeid, first_college_any_opeid, 
#          first_college_4yr_opeid, first_college_2yr_opeid, 
#          days_enrolled) %>% 
#   filter(sid == 13653) %>% as.data.frame
# 




# // If a student started at multiple colleges of the same type on the same date, indicate this by replacing

# // these values with a dummy value (“>1”) for processing.
# replace temp_first_college_`var’_`type’ = “>1” if nvals_first_college_`var’_`type’ > 1 & nvals_first_college_`var’_`type’
# !=.


```



** Step 3: Drop any unneeded variables, and save the file **
```{r}
# // 1. Drop the unneeded variables
# drop temp* nvals* days_enrolled
# // 2. Save the current file as Student_NSC_Enrollment_Indicators
# save “${clean}\Student_NSC_Enrollment_Indicators.dta”, replace
```

